---
title: "Batch effects exploration"
author: "Kushal K Dey"
date: "February 4, 2016"
output: 
  html_document:
    css: floating-toc.css
    toc: true
---

## Overview

In this script, we explore the effectiveness of removing batch effects by the `BatchCorrectedCounts()` function of the package `CountClust`. We consider three simulation scenarios and then fit the topic model before and after removing the batch effects. 

```{r echo=TRUE, eval=TRUE}
library(CountClust)
library(maptpx)
```


## Simulation Scenario 1

We present the simulation mechanism for scenario 1.

```{r echo=TRUE, eval=TRUE}
K=4;
G=100;
N=200;
alpha_true=matrix(rnorm((K)*G,0.5,1),nrow=(K)); ### the matrix of fixed effects
Label.Batch=c(rep(1,N/2),rep(2,N/2));
B=max(Label.Batch);
sigmab_true=2;
beta_true=matrix(0,B,G);       ###  the matrix of the random effect
for(g in 1:G)
{
  beta_true[,g]=rnorm(B,mean=0,sd=sigmab_true);
}
library(gtools)
T=10;
omega_true=matrix(rbind(rdirichlet(T*4,c(3,4,2,6)),rdirichlet(T*4,c(1,4,6,3)),
                        rdirichlet(T*4,c(4,1,2,2)),rdirichlet(T*4,c(2,6,3,2)),
                        rdirichlet(T*4,c(3,3,5,4))), nrow=N);

over_dis=0.5;
noise_true=matrix(0,N,G);
for(n in 1:N)
{
  noise_true[n,]=over_dis*rnorm(G,0,1);
}

###  generating the table
read_counts=matrix(0,N,G);
for(n in 1:N)
{
  for(g in 1:G)
  {
    mean=exp(omega_true[n,]%*%alpha_true[,g] +beta_true[Label.Batch[n],g]+noise_true[n,g]);
    read_counts[n,g]=rpois(1,mean);
  }
}


```

### True Structure Plot

```{r echo=TRUE, eval=TRUE}
barplot(t(omega_true),col=2:(K+1),axisnames=F,space=0,border=NA,main="",las=1,ylim=c(0,1),cex.axis=1.5,cex.main=1.4)
title(main=paste("Structure Plot of initial chosen topic proportions,k=",K))
```

### Topic model (Without Batch Correction)

```{r echo=TRUE, eval=TRUE, warning=FALSE, message=FALSE}
Topic_clus <- topics(read_counts, K, tol=0.01);
docweights <- Topic_clus$omega;
```

### Estimated Structure plot (Without Batch correction)

```{r echo=TRUE, eval=TRUE}
barplot(t(docweights),col=2:(K+1),axisnames=F,space=0,border=NA,main="",las=1,ylim=c(0,1),cex.axis=1.5,cex.main=1.4)
title(main=paste("Structure Plot of initial chosen topic proportions,k=",K))
```

We perform the batch correction

### Batch corrected topic model fit

```{r echo=TRUE, eval=TRUE}
batch_counts <- BatchCorrectedCounts(read_counts, batch_lab = Label.Batch);

Topic_clus_2 <- topics(batch_counts, K, tol=0.01);
docweights_2 <- Topic_clus_2$omega;
```

### Estimated Structure Plot (Batch correction)

```{r echo=TRUE, eval=TRUE}
library(permute);
library("BioPhysConnectoR");
perm_set=rbind(1:K,allPerms(1:K));
diff=array(0,dim(perm_set)[1]);
for (p in 1:dim(perm_set)[1])
{
  temp=docweights_2[,perm_set[p,]];
  diff[p]=fnorm(temp,omega_true);
}

p_star=which(diff==min(diff));
docweights_2=docweights_2[,perm_set[p_star,]];

barplot(t(docweights_2),col=2:(K+1),axisnames=F,space=0,border=NA,main="",las=1,ylim=c(0,1),cex.axis=1.5,cex.main=1.4)
title(main=paste("Structure Plot of initial chosen topic proportions,k=",K))
```


## Simulation Scenario 2

We now present simulation scenario 2

```{r echo=TRUE, eval=TRUE}
K=4;
G=100;
N=200;
alpha_true=matrix(rnorm((K)*G,0.5,1),nrow=(K)); ### the matrix of fixed effects
Label.Batch=c(rep(1,N/2),rep(2,N/2));
B=max(Label.Batch);
sigmab_true=2;
beta_true=matrix(0,B,G);       ###  the matrix of the random effect
for(g in 1:G)
{
  beta_true[,g]=rnorm(B,mean=0,sd=sigmab_true);
}
library(gtools)
T=10;
omega_true=matrix(rbind(rdirichlet(T*4,c(3,4,2,6)),rdirichlet(T*4,c(1,4,6,3)),
                        rdirichlet(T*4,c(4,1,2,2)),rdirichlet(T*4,c(2,6,3,2)),
                        rdirichlet(T*4,c(3,3,5,4))), nrow=N);

over_dis=0.5;
noise_true=matrix(0,N,G);
for(n in 1:N)
{
  noise_true[n,]=over_dis*rnorm(G,0,1);
}

###  generating the table
read_counts=matrix(0,N,G);
for(n in 1:N)
{
  for(g in 1:G)
  {
    mean=omega_true[n,]%*%exp(alpha_true[,g] +beta_true[Label.Batch[n],g]+noise_true[n,g]);
    read_counts[n,g]=rpois(1,mean);
  }
}


```

### True Structure Plot

```{r echo=TRUE, eval=TRUE}
barplot(t(omega_true),col=2:(K+1),axisnames=F,space=0,border=NA,main="",las=1,ylim=c(0,1),cex.axis=1.5,cex.main=1.4)
title(main=paste("Structure Plot of initial chosen topic proportions,k=",K))
```

### Topic model (Without Batch Correction)

```{r echo=TRUE, eval=TRUE, warning=FALSE, message=FALSE}
Topic_clus <- topics(read_counts, K, tol=0.01);
docweights <- Topic_clus$omega;
```

### Estimated Structure plot (Without Batch correction)

```{r echo=TRUE, eval=TRUE}
barplot(t(docweights),col=2:(K+1),axisnames=F,space=0,border=NA,main="",las=1,ylim=c(0,1),cex.axis=1.5,cex.main=1.4)
title(main=paste("Structure Plot of initial chosen topic proportions,k=",K))
```

We perform the batch correction

### Batch corrected topic model fit

```{r echo=TRUE, eval=TRUE}
batch_counts <- BatchCorrectedCounts(read_counts, batch_lab = Label.Batch);

Topic_clus_2 <- topics(batch_counts, K, tol=0.01);
docweights_2 <- Topic_clus_2$omega;
```

### Estimated Structure Plot (Batch correction)

```{r echo=TRUE, eval=TRUE}
library(permute);
library("BioPhysConnectoR");
perm_set=rbind(1:K,allPerms(1:K));
diff=array(0,dim(perm_set)[1]);
for (p in 1:dim(perm_set)[1])
{
  temp=docweights_2[,perm_set[p,]];
  diff[p]=fnorm(temp,omega_true);
}

p_star=which(diff==min(diff));
docweights_2=docweights_2[,perm_set[p_star,]];

barplot(t(docweights_2),col=2:(K+1),axisnames=F,space=0,border=NA,main="",las=1,ylim=c(0,1),cex.axis=1.5,cex.main=1.4)
title(main=paste("Structure Plot of initial chosen topic proportions,k=",K))
```

## Simualtion Scenario 3

We present the simulation scenario 3.

```{r eho=TRUE, eval=TRUE}
K=4;
G=100;
N=200;

Label.Batch=c(rep(1,N/2),rep(2,N/2));
B=max(Label.Batch);
sigmab_true=2;
beta_true=matrix(0,B,G);       ###  the matrix of the random effect
for(g in 1:G)
{
  beta_true[,g]=rnorm(B,mean=0,sd=sigmab_true);
}
library(gtools)
T=10;
omega_true=matrix(rbind(rdirichlet(T*4,c(3,4,2,6)),rdirichlet(T*4,c(1,4,6,3)),
                        rdirichlet(T*4,c(4,1,2,2)),rdirichlet(T*4,c(2,6,3,2)),
                        rdirichlet(T*4,c(3,3,5,4))), nrow=N);

theta_b_true <- rbind(rdirichlet(2, c(0.2, 0.2, rep(0.6/G,G-2))),
                           rdirichlet(2, c(rep(0.5/G,G-3),0.3,0.1,0.1)),
                           rdirichlet(2, c(rep(0.4/G, G/2 -2),0.2,0.1,rep(0.3/G, G/2))),
                           rdirichlet(2, c(rep(0.6/G,G-2), 0.2, 0.2)));

read_counts <- t(do.call(cbind,lapply(1:dim(omega_true)[1], function(x)
                                                        {
                                                            if(Label.Batch[x]==1)
                                                              out <- rmultinom(1,1000,prob=omega_true[x,]%*%theta_b_true[c(2,4,6,8),]);
                                                            if(Label.Batch[x]==2)
                                                              out <- rmultinom(1,1000,prob=omega_true[x,]%*%theta_b_true[c(1,3,5,7),]);
                                                            return(out)
                                                             }  )));
```

### True Structure Plot

```{r echo=TRUE, eval=TRUE}
barplot(t(omega_true),col=2:(K+1),axisnames=F,space=0,border=NA,main="",las=1,ylim=c(0,1),cex.axis=1.5,cex.main=1.4)
title(main=paste("Structure Plot of initial chosen topic proportions,k=",K))
```

### Topic model (Without Batch Correction)

```{r echo=TRUE, eval=TRUE, warning=FALSE, message=FALSE}
Topic_clus <- topics(read_counts, K, tol=0.01);
docweights <- Topic_clus$omega;
```

### Estimated Structure plot (Without Batch correction)

```{r echo=TRUE, eval=TRUE}
barplot(t(docweights),col=2:(K+1),axisnames=F,space=0,border=NA,main="",las=1,ylim=c(0,1),cex.axis=1.5,cex.main=1.4)
title(main=paste("Structure Plot of initial chosen topic proportions,k=",K))
```

We perform the batch correction

### Batch corrected topic model fit

```{r echo=TRUE, eval=TRUE}
batch_counts <- BatchCorrectedCounts(read_counts, batch_lab = Label.Batch);

Topic_clus_2 <- topics(batch_counts, K, tol=0.01);
docweights_2 <- Topic_clus_2$omega;
```

### Estimated Structure Plot (Batch correction)

```{r echo=TRUE, eval=TRUE}
library(permute);
library("BioPhysConnectoR");
perm_set=rbind(1:K,allPerms(1:K));
diff=array(0,dim(perm_set)[1]);
for (p in 1:dim(perm_set)[1])
{
  temp=docweights_2[,perm_set[p,]];
  diff[p]=fnorm(temp,omega_true);
}

p_star=which(diff==min(diff));
docweights_2=docweights_2[,perm_set[p_star,]];

barplot(t(docweights_2),col=2:(K+1),axisnames=F,space=0,border=NA,main="",las=1,ylim=c(0,1),cex.axis=1.5,cex.main=1.4)
title(main=paste("Structure Plot of initial chosen topic proportions,k=",K))
```

## Conclusions

It seems that although the batch correction does improve on model fitting and seem to remove the batch effects problem, but the model fit after batch correction and corresponding Structure plot representation does not correspond very accurately to the true Structure plot representation.

