---
title: 'Gene Annotation : GTEx'
author: "Kushal K Dey"
date: "July 5, 2015"
output: 
  html_document:
    css: floating-toc.css
    toc: true
---

```{r packages, echo=FALSE, eval=TRUE, warning=FALSE, message=FALSE}
library(maptpx)
library(qtlcharts)
library(data.table)
library(mygene)
library(CountClust)
library(knitr)
```

## Overview

In this script, we give the gene annotations of the genes that seem to vary greatly between the clusters we get. This would  be indicative of the markers that are driving the different clusters. If we can find that for a cluster which is mainly represented in a particular tissue type, if the genes significantly differentially expressed in that cluster are indeed related to the tissue in terms of its annotation, then we can say that the clustering makes biological sense. 

## Extracting top driving genes 

```{r gene_annotation, echo=TRUE, eval=TRUE}

topics_theta <- data.frame(fread("../external_data/GTEX_V6/admix_out_GTEX_V6/theta_cis_genes_0_1_2.txt"))[,-1];

top_features <- ExtractTopFeatures(topics_theta, top_features=100, method="poisson", options="min");

gene_names <- as.vector(as.matrix(read.table("../external_data/GTEX_V6/gene_names_GTEX_V6.txt")))
gene_names <- substring(gene_names,1,15);
xli  <-  gene_names;
gene_list <- do.call(rbind, lapply(1:dim(top_features)[1], function(x) gene_names[top_features[x,]]))

```

## Cluster 1 Annotations

```{r echo=TRUE, eval=TRUE}
out <- mygene::queryMany(gene_list[1,],  scopes="ensembl.gene", fields=c("name", "summary"), species="human");
kable(as.data.frame(out))
```

## Cluster 2 Annotations

```{r echo=TRUE, eval=TRUE}
out <- mygene::queryMany(gene_list[2,],  scopes="ensembl.gene", fields=c("name", "summary"), species="human");
kable(as.data.frame(out))
```

## Cluster 3 Annotations

```{r echo=TRUE, eval=TRUE}
out <- mygene::queryMany(gene_list[3,],  scopes="ensembl.gene", fields=c("name", "summary"), species="human");
kable(as.data.frame(out))
```

## Cluster 4 Annotations

```{r echo=TRUE, eval=TRUE}
out <- mygene::queryMany(gene_list[4,],  scopes="ensembl.gene", fields=c("name", "summary"), species="human");
kable(as.data.frame(out))
```

## Cluster 5 Annotations

```{r echo=TRUE, eval=TRUE}
out <- mygene::queryMany(gene_list[5,],  scopes="ensembl.gene", fields=c("name", "summary"), species="human");
kable(as.data.frame(out))
```

## Cluster 6 Annotations

```{r echo=TRUE, eval=TRUE}
out <- mygene::queryMany(gene_list[6,],  scopes="ensembl.gene", fields=c("name", "summary"), species="human");
kable(as.data.frame(out))
```

## Cluster 7 Annotations

```{r echo=TRUE, eval=TRUE}
out <- mygene::queryMany(gene_list[7,],  scopes="ensembl.gene", fields=c("name", "summary"), species="human");
kable(as.data.frame(out))
```

## Cluster 8 Annotations

```{r echo=TRUE, eval=TRUE}
out <- mygene::queryMany(gene_list[8,],  scopes="ensembl.gene", fields=c("name", "summary"), species="human");
kable(as.data.frame(out))
```

## Cluster 9 Annotations

```{r echo=TRUE, eval=TRUE}
out <- mygene::queryMany(gene_list[9,],  scopes="ensembl.gene", fields=c("name", "summary"), species="human");
kable(as.data.frame(out))
```

## Cluster 10 Annotations

```{r echo=TRUE, eval=TRUE}
out <- mygene::queryMany(gene_list[10,],  scopes="ensembl.gene", fields=c("name", "summary"), species="human");
kable(as.data.frame(out))
```

## Cluster 11 Annotations

```{r echo=TRUE, eval=TRUE}
out <- mygene::queryMany(gene_list[11,],  scopes="ensembl.gene", fields=c("name", "summary"), species="human");
kable(as.data.frame(out))
```

## Cluster 12 Annotations

```{r echo=TRUE, eval=TRUE}
out <- mygene::queryMany(gene_list[12,],  scopes="ensembl.gene", fields=c("name", "summary"), species="human");
kable(as.data.frame(out))
```

## Cluster 13 Annotations

```{r echo=TRUE, eval=TRUE}
out <- mygene::queryMany(gene_list[13,],  scopes="ensembl.gene", fields=c("name", "summary"), species="human");
kable(as.data.frame(out))
```

## Cluster 14 Annotations

```{r echo=TRUE, eval=TRUE}
out <- mygene::queryMany(gene_list[14,],  scopes="ensembl.gene", fields=c("name", "summary"), species="human");
kable(as.data.frame(out))
```

## Cluster 15 Annotations

```{r echo=TRUE, eval=TRUE}
out <- mygene::queryMany(gene_list[15,],  scopes="ensembl.gene", fields=c("name", "summary"), species="human");
kable(as.data.frame(out))
```


