---
title: 'Gene Annotation : GTEx'
author: "Kushal K Dey"
date: "July 5, 2015"
output: 
  html_document:
    css: floating-toc.css
    toc: true
---

```{r packages, echo=FALSE, eval=TRUE, warning=FALSE, message=FALSE}
library(maptpx)
library(qtlcharts)
library(data.table)
library(mygene)

```

## Overview

In this script, we give the gene annotations of the genes that seem to vary greatly between the clusters we get. This would  be indicative of the markers that are driving the different clusters. If we can find that for a cluster which is mainly represented in a particular tissue type, if the genes significantly differentially expressed in that cluster are indeed related to the tissue in terms of its annotation, then we can say that the clustering makes biological sense. 

## KL divergence scheme

```{r gene_annotation, echo=TRUE, eval=TRUE}

topics_theta <- data.frame(fread("../external_data/GTEX_V6/admix_out_GTEX_V6/theta_cis_genes_0_1_2.txt"))[,-1];

sqrt_zscore <- t(apply(topics_theta, 1, function(x)
                                      {
                                          y=sqrt(x);
                                          return((y-mean(y))/sd(y));
                                      }))

log_zscore <- t(apply(topics_theta, 1, function(x)
                                      {
                                          y=log((x)/(1-x));
                                          return((y-mean(y))/sd(y));
                                      }))
K=dim(topics_theta)[2];

KL_score_poisson <- lapply(1:K, function(n) 
                                {
                                    out <- t(apply(topics_theta, 1, function(x)
                                                                    {
                                                                      y=x[n] *log(x[n]/x) + x - x[n];
                                                                      return(y)
                                                                     }));
                                    return(out)
                                })
                   
KL_score_bernoulli <- lapply(1:K, function(n) 
                                {
                                    out <- t(apply(topics_theta, 1, function(x)
                                                                    {
                                                                      y=x[n] *log(x[n]/x) + (1 - x[n])*log((1-x[n])/(1-x));
                                                                      return(y)
                                                                     }));
                                    return(out)
                                })
                   
```

We used four divergence conditions 

- Square root z - transform 

- Log z- transform

- Bernoulli K-L divergence

- Poisson K-L divergence 

Out of these conditions, Square root z- transform and Log z-transform gave close results. On the other hand, Bernoulli K-L divergence and Poisson K-L divergence gave practically identical results (almost). So, we focused on the Poisson K-L divergence and the Square root Z- transform instead. 

```{r echo=FALSE, eval=TRUE}

indices_mat_poisson=matrix(0,dim(topics_theta)[2],10);

for(k in 1:dim(topics_theta)[2])
{
  temp_mat <- KL_score_poisson[[k]][,-k];
  vec <- apply(temp_mat, 1, function(x) min(x))
  indices_mat_poisson[k,] = order(vec, decreasing = TRUE)[1:10]
}

#gene_names <- substring(gene_snp_names_gtex[,2],1,15);

gene_names <- as.vector(as.matrix(read.table("../external_data/GTEX_V6/gene_names_GTEX_V6.txt")))
gene_names <- substring(gene_names,1,15);
gene_names_mat_poisson = matrix(gene_names[as.vector(indices_mat_poisson)],nrow=dim(topics_theta)[2]);

color=c("red","blue","cornflowerblue","black","cyan","darkblue",
        "brown4","burlywood","darkgoldenrod1","darkgray","deepskyblue","darkkhaki",
        "firebrick","darkorchid","hotpink","green","magenta","yellow", "azure1","azure4");



```

## Gene Annotations (mygene package)

Use **mygene** R package to get information about all the cis- genes used in the GTEx data.
It creates a database of the cis genes with summary level information.

```{r, echo=FALSE, eval=TRUE}

xli  <-  gene_names;
out <- queryMany(xli, scopes="ensembl.gene", fields=c("name", "summary"), species="human");

#write.table(out, "/Users/kushal/Documents/gtex-viz/gtex.Kushal/data/cis_gene_complete_summary_gtex.txt");

#write.table(out, "/Users/kushal/Documents/gtex-viz/gtex.Kushal/data/reads_gene_complete_summary_gtex.txt");


out <- read.table("/Users/kushal/Documents/gtex-viz/gtex.Kushal/data/reads_gene_complete_summary_gtex.txt");

par(mfrow=c(1,1))
barplot(rep(1,15),col=color[1:15])
```


## Structure plot (K=15)

```{r}

omega <- data.frame(fread("../external_data/GTEX_V6/admix_out_GTEX_V6/omega_cis_genes_0_1_2.txt"))[,-1];

samples_id=read.table("../external_data/GTEX_V6/samples_id.txt")[,3];
docweights_per_tissue_mean <- apply(omega,2,function(x) tapply(x,samples_id,mean));

ordering=heatmap(docweights_per_tissue_mean)$rowInd;

unique_samples_id_ordered = unique(samples_id)[ordering];
clus_ordered =unlist(lapply(1:53, function(x) which(samples_id == unique_samples_id_ordered[x])));
samples_id_ordered = samples_id[clus_ordered];

docweights_ordering = omega[clus_ordered,];

par(mar=c(14,2.5,1,1))
barplot(t(docweights_ordering),col=color[1:K],axisnames=F,space=0,border=NA,main=paste("No. of clusters=",K),las=1,ylim=c(0,1),cex.axis=1.2,cex.main=1.4)
  
labels = match(unique(samples_id_ordered), samples_id_ordered);
#abline(v=labels)
  
labels_low=labels;
labels_up=c(labels[2:length(labels)],dim(docweights_ordering)[1]);
mid_point=labels_low +0.5*(labels_up-labels_low);
axis(1,at=mid_point, unique(samples_id_ordered),las=2);

  

```

## Color bars of clusters

The colors of the bars corresponding to different cluster numbers are as follows (bars arranged in sequence of the cluster number)

```{r echo=TRUE, eval=TRUE}
barplot(rep(1,15),col=color[1:15])
```

## Gene summaries individual clusters 

### red cluster (cluster 1)

```{r echo=FALSE, eval=TRUE}

clus <- 1
gene_names_mat_poisson[clus,]
indices <- sapply(1:length(gene_names_mat_poisson[clus,]), function(n) grep(gene_names_mat_poisson[clus,n],gene_names))
indices_to_consider <- which(apply(topics_theta[indices,],1, function(x) which.max(x))==clus);
new_genes <- gene_names_mat_poisson[clus,indices_to_consider];
lapply(1:length(new_genes), function(n) out[grep(as.character(new_genes[n]), out$query),])
lapply(1:length(gene_names_mat_poisson[clus,]
), function(n) out[grep(as.character(gene_names_mat_poisson[clus,n]
), out$query), ])

```


### blue cluster (cluster 2)

```{r echo=FALSE, eval=TRUE}

clus <- 2
gene_names_mat_poisson[clus,]
indices <- sapply(1:length(gene_names_mat_poisson[clus,]), function(n) grep(gene_names_mat_poisson[clus,n],gene_names))
indices_to_consider <- which(apply(topics_theta[indices,],1, function(x) which.max(x))==clus);
new_genes <- gene_names_mat_poisson[clus,indices_to_consider];
lapply(1:length(new_genes), function(n) out[grep(as.character(new_genes[n]), out$query),])
lapply(1:length(gene_names_mat_poisson[clus,]
), function(n) out[grep(as.character(gene_names_mat_poisson[clus,n]
), out$query), ])

```


### shallow blue cluster (cluster 3)

```{r echo=FALSE, eval=TRUE}

clus <- 3
gene_names_mat_poisson[clus,]
indices <- sapply(1:length(gene_names_mat_poisson[clus,]), function(n) grep(gene_names_mat_poisson[clus,n],gene_names))
indices_to_consider <- which(apply(topics_theta[indices,],1, function(x) which.max(x))==clus);
new_genes <- gene_names_mat_poisson[clus,indices_to_consider];
lapply(1:length(new_genes), function(n) out[grep(as.character(new_genes[n]), out$query),])
lapply(1:length(gene_names_mat_poisson[clus,]
), function(n) out[grep(as.character(gene_names_mat_poisson[clus,n]
), out$query), ])

```

### black cluster (cluster 4)

```{r echo=FALSE, eval=TRUE}

clus <- 4
gene_names_mat_poisson[clus,]
indices <- sapply(1:length(gene_names_mat_poisson[clus,]), function(n) grep(gene_names_mat_poisson[clus,n],gene_names))
indices_to_consider <- which(apply(topics_theta[indices,],1, function(x) which.max(x))==clus);
new_genes <- gene_names_mat_poisson[clus,indices_to_consider];
lapply(1:length(new_genes), function(n) out[grep(as.character(new_genes[n]), out$query),])
lapply(1:length(gene_names_mat_poisson[clus,]
), function(n) out[grep(as.character(gene_names_mat_poisson[clus,n]
), out$query), ])

```

### light blue cluster (cluster 5)

```{r echo=FALSE, eval=TRUE}

clus <- 5
gene_names_mat_poisson[clus,]
indices <- sapply(1:length(gene_names_mat_poisson[clus,]), function(n) grep(gene_names_mat_poisson[clus,n],gene_names))
indices_to_consider <- which(apply(topics_theta[indices,],1, function(x) which.max(x))==clus);
new_genes <- gene_names_mat_poisson[clus,indices_to_consider];
lapply(1:length(new_genes), function(n) out[grep(as.character(new_genes[n]), out$query),])
lapply(1:length(gene_names_mat_poisson[clus,]
), function(n) out[grep(as.character(gene_names_mat_poisson[clus,n]
), out$query), ])

```

### deep blue cluster (cluster 6)

```{r echo=FALSE, eval=TRUE}

clus <- 6
gene_names_mat_poisson[clus,]
indices <- sapply(1:length(gene_names_mat_poisson[clus,]), function(n) grep(gene_names_mat_poisson[clus,n],gene_names))
indices_to_consider <- which(apply(topics_theta[indices,],1, function(x) which.max(x))==clus);
new_genes <- gene_names_mat_poisson[clus,indices_to_consider];
lapply(1:length(new_genes), function(n) out[grep(as.character(new_genes[n]), out$query),])
lapply(1:length(gene_names_mat_poisson[clus,]
), function(n) out[grep(as.character(gene_names_mat_poisson[clus,n]
), out$query), ])

```


### dark brown cluster (cluster 7)

```{r echo=FALSE, eval=TRUE}

clus <- 7
gene_names_mat_poisson[clus,]
indices <- sapply(1:length(gene_names_mat_poisson[clus,]), function(n) grep(gene_names_mat_poisson[clus,n],gene_names))
indices_to_consider <- which(apply(topics_theta[indices,],1, function(x) which.max(x))==clus);
new_genes <- gene_names_mat_poisson[clus,indices_to_consider];
lapply(1:length(new_genes), function(n) out[grep(as.character(new_genes[n]), out$query),])
lapply(1:length(gene_names_mat_poisson[clus,]
), function(n) out[grep(as.character(gene_names_mat_poisson[clus,n]
), out$query), ])

```

### shallow yellow cluster (cluster 8)

```{r echo=FALSE, eval=TRUE}

clus <- 8
gene_names_mat_poisson[clus,]
indices <- sapply(1:length(gene_names_mat_poisson[clus,]), function(n) grep(gene_names_mat_poisson[clus,n],gene_names))
indices_to_consider <- which(apply(topics_theta[indices,],1, function(x) which.max(x))==clus);
new_genes <- gene_names_mat_poisson[clus,indices_to_consider];
lapply(1:length(new_genes), function(n) out[grep(as.character(new_genes[n]), out$query),])
lapply(1:length(gene_names_mat_poisson[clus,]
), function(n) out[grep(as.character(gene_names_mat_poisson[clus,n]
), out$query), ])

```


### yellow cluster (cluster 9)

```{r echo=FALSE, eval=TRUE}

clus <- 9
gene_names_mat_poisson[clus,]
indices <- sapply(1:length(gene_names_mat_poisson[clus,]), function(n) grep(gene_names_mat_poisson[clus,n],gene_names))
indices_to_consider <- which(apply(topics_theta[indices,],1, function(x) which.max(x))==clus);
new_genes <- gene_names_mat_poisson[clus,indices_to_consider];
lapply(1:length(new_genes), function(n) out[grep(as.character(new_genes[n]), out$query),])
lapply(1:length(gene_names_mat_poisson[clus,]
), function(n) out[grep(as.character(gene_names_mat_poisson[clus,n]
), out$query), ])

```


### grey cluster (cluster 10)

```{r echo=FALSE, eval=TRUE}

clus <- 10
gene_names_mat_poisson[clus,]
indices <- sapply(1:length(gene_names_mat_poisson[clus,]), function(n) grep(gene_names_mat_poisson[clus,n],gene_names))
indices_to_consider <- which(apply(topics_theta[indices,],1, function(x) which.max(x))==clus);
new_genes <- gene_names_mat_poisson[clus,indices_to_consider];
lapply(1:length(new_genes), function(n) out[grep(as.character(new_genes[n]), out$query),])
lapply(1:length(gene_names_mat_poisson[clus,]
), function(n) out[grep(as.character(gene_names_mat_poisson[clus,n]
), out$query), ])

```

### cyan cluster (cluster 11)

```{r echo=FALSE, eval=TRUE}

clus <- 11
gene_names_mat_poisson[clus,]
indices <- sapply(1:length(gene_names_mat_poisson[clus,]), function(n) grep(gene_names_mat_poisson[clus,n],gene_names))
indices_to_consider <- which(apply(topics_theta[indices,],1, function(x) which.max(x))==clus);
new_genes <- gene_names_mat_poisson[clus,indices_to_consider];
lapply(1:length(new_genes), function(n) out[grep(as.character(new_genes[n]), out$query),])
lapply(1:length(gene_names_mat_poisson[clus,]
), function(n) out[grep(as.character(gene_names_mat_poisson[clus,n]
), out$query), ])

```

### shallow green cluster (cluster 12)

```{r echo=FALSE, eval=TRUE}

clus <- 12
gene_names_mat_poisson[clus,]
indices <- sapply(1:length(gene_names_mat_poisson[clus,]), function(n) grep(gene_names_mat_poisson[clus,n],gene_names))
indices_to_consider <- which(apply(topics_theta[indices,],1, function(x) which.max(x))==clus);
new_genes <- gene_names_mat_poisson[clus,indices_to_consider];
lapply(1:length(new_genes), function(n) out[grep(as.character(new_genes[n]), out$query),])
lapply(1:length(gene_names_mat_poisson[clus,]
), function(n) out[grep(as.character(gene_names_mat_poisson[clus,n]
), out$query), ])

```


### light brown cluster (cluster 13)

```{r echo=FALSE, eval=TRUE}

clus <- 13
gene_names_mat_poisson[clus,]
indices <- sapply(1:length(gene_names_mat_poisson[clus,]), function(n) grep(gene_names_mat_poisson[clus,n],gene_names))
indices_to_consider <- which(apply(topics_theta[indices,],1, function(x) which.max(x))==clus);
new_genes <- gene_names_mat_poisson[clus,indices_to_consider];
lapply(1:length(new_genes), function(n) out[grep(as.character(new_genes[n]), out$query),])
lapply(1:length(gene_names_mat_poisson[clus,]
), function(n) out[grep(as.character(gene_names_mat_poisson[clus,n]
), out$query), ])

```


### violet  cluster (cluster 14)

```{r echo=FALSE, eval=TRUE}

clus <- 14
gene_names_mat_poisson[clus,]
indices <- sapply(1:length(gene_names_mat_poisson[clus,]), function(n) grep(gene_names_mat_poisson[clus,n],gene_names))
indices_to_consider <- which(apply(topics_theta[indices,],1, function(x) which.max(x))==clus);
new_genes <- gene_names_mat_poisson[clus,indices_to_consider];
lapply(1:length(new_genes), function(n) out[grep(as.character(new_genes[n]), out$query),])
lapply(1:length(gene_names_mat_poisson[clus,]
), function(n) out[grep(as.character(gene_names_mat_poisson[clus,n]
), out$query), ])

```


### salmon  cluster (cluster 15)

```{r echo=FALSE, eval=TRUE}

clus <- 15
gene_names_mat_poisson[clus,]
indices <- sapply(1:length(gene_names_mat_poisson[clus,]), function(n) grep(gene_names_mat_poisson[clus,n],gene_names))
indices_to_consider <- which(apply(topics_theta[indices,],1, function(x) which.max(x))==clus);
new_genes <- gene_names_mat_poisson[clus,indices_to_consider];
lapply(1:length(new_genes), function(n) out[grep(as.character(new_genes[n]), out$query),])
lapply(1:length(gene_names_mat_poisson[clus,]
), function(n) out[grep(as.character(gene_names_mat_poisson[clus,n]
), out$query), ])

```


