---
title: 'GTEx: sort tissues by germ layers of origin'
author: "Joyce Hsiao"
date: "2016-09-21"
output: 
  html_document:
    css: floating-toc.css
    toc: true
---


```{r chunk-options, include=FALSE}
source("chunk-options.R")
```


## Objective

In order to aid the biological interpretation of GTEx tissue clustering, we made a dendrogram of the tissues based on the multivariate distance matrix of admixtures. Specifically, we take the GoM fitting of 20 clusters, compute the Friedman-Rafsky test statistic comparing the admixtures between any two tissues, then use make a dendrogram of the resulting distance matrix. We use this dendrogram to gain insight into tissue variation in the Structure plot.


## Setting up

Load packages

```{r packages, echo=FALSE, eval=TRUE, warning=FALSE, message=FALSE}
#library(maptpx)
library(ggplot2)
library(testit)
library(data.table)
library(mygene)
library(knitr)
library(flowMap)
```


Import GoM fit

```{r}
fit <- get(load("../rdas/gtexv6fit.k.20.master.rda"))
omega <- fit$omega
```

Load sample labels

```{r}
sample_labels <- read.table("../rdas/samples-id-for-paper.txt",
                            header = TRUE, sep = "\t",
                            stringsAsFactors = FALSE)
sample_labels$SMTSD <- factor(sample_labels$SMTSD)
sample_labels_list <- levels(sample_labels$SMTSD)
```


## Multivariate distance between tissues

Make an empty distance matrix

```{r}
distmat_fr <- matrix(0, nrow = length(sample_labels_list),
                        ncol = length(sample_labels_list))
```

Make an index matrix for upper triangular matrix

```{r}
distmat_upper_index <- which( upper.tri(distmat_fr, diag = FALSE),
                              arr.ind = TRUE)
```

Fill in the distance matrix. This requres 1,378 calculations of FR distances. 

```{r}
rda_name <- "../rdas/gtex-tissue-distance.rda"
if (file.exists(rda_name)) {
    load(rda_name)
} else {
    message("The GTEx distance matrix does not exist. Compute the distances now.")
    for (i in 1:nrow(distmat_upper_index)) {
        tissue1_index <- distmat_upper_index[i,1]
        tissue2_index <- distmat_upper_index[i,2]
        distmat_fr[distmat_upper_index[i,1], distmat_upper_index[i,2]] <- 
            getFR(omega[sample_labels$SMTSD == sample_labels_list[tissue1_index], ],
                  omega[sample_labels$SMTSD == sample_labels_list[tissue2_index], ])$ww
    }
    
    rownames(distmat_fr) <- sample_labels_list
    colnames(distmat_fr) <- sample_labels_list

    # Make distmat_fr into a distance matrix
    # library(Matrix)
    distmat_fr2 <- Matrix::sparseMatrix(
            i=distmat_upper_index[,1],
            j=distmat_upper_index[,2],
            x=distmat_fr[upper.tri(distmat_fr)],
            symmetric=TRUE,    
            dims=c(nrow(distmat_fr),nrow(distmat_fr)))

    distmat <- -1*distmat_fr2
    distmat <- as.dist(distmat)
    
    save(distmat, file = "../rdas/gtex-tissue-distance.rda")
}
```


## Make dendrogram

```{r gtex-gom-clusters-dendrogram, fig.height=9, fig.width=9}
colors <- c(seq(-2,10,length=120),seq(10,30,length=200))
colors <- colors[!duplicated(colors)]
my_palette <- colorRampPalette(
    c("blue", "yellow", "orange", "orangered", "red4"))(n = length(colors)-1)
gplots::heatmap.2(as.matrix(distmat), 
                  breaks = colors,
                  col = my_palette,
                  labRow = sample_labels_list,
                  trace = "none", dendrogram = "row",
                  symm=F,symkey=F,symbreaks=T, scale="none")
```


## Session information

```{r, echo = FALSE, include = FALSE}
sessionInfo()
```

