---
title: "Meta topics analysis GTEx brain V6"
author: "Kushal K Dey"
date: "April 15, 2016"
output: html_document
---

## Overview

We have used topic model or admixture to cluster GTEx V6 Brain samples. But motivated by the Deep Poisson Factor Modeling by Lawrence Carin's group, we may try to see if there is information to be had in looking at the metatopics for the GTEx V6 Brain samples. The idea is to see if there is any merit into looking at the metatopics and whether they are giving any more intuition than the Admixture plot.

```{r echo=TRUE, eval=TRUE}
omega_brain <- read.table("../external_data/GTEX_V6/admix_out_GTEX_V6/omega_cis_genes_brain.txt");
theta_brain <- read.table("../external_data/GTEX_V6/admix_out_GTEX_V6/theta_cis_genes_brain.txt");

data <- data.frame(fread('../external_data/GTEX_V6/cis_gene_expression.txt'));
matdata <- data[,-(1:2)];
sample_labels <- read.table("../external_data/GTEX_V6/samples_id.txt",
                            header = TRUE, sep = " ",
                            stringsAsFactors = FALSE)
brain_labels <- sample_labels[grep("Brain", sample_labels[,3]), 3]

brain_data <- matdata[,grep("Brain", sample_labels[,3])];
libsize_brain <- colSums(brain_data);

lambda_brain <- sweep(omega_brain, 1, libsize_brain, "*");

z_level_1 <- array(0,c(dim(lambda_brain)[1], dim(theta_brain)[1], dim(lambda_brain)[2]));
for(k in 1:dim(theta_brain)[2]){
  z_level_1[,,k] <- lambda_brain[,k]%*%t(theta_brain[,k])
}

z_level_1_counts <- floor(apply(z_level_1, c(1,3), sum));

```

We apply topic model with $K2=3$ on this matrix.

```{r}
topic_clus <- maptpx::topics(z_level_1_counts, K=3, tol=0.1);
```

```{r}
rownames(omega_brain) <- paste0("X", 1:length(brain_labels))
annotation <- data.frame(
    sample_id = paste0("X", 1:length(brain_labels)),
    tissue_label = factor(brain_labels,
                          levels = rev(c("Brain - Cerebellar Hemisphere",
                                     "Brain - Cerebellum",
                                     "Brain - Spinal cord (cervical c-1)",
                                     "Brain - Anterior cingulate cortex (BA24)",
                                     "Brain - Frontal Cortex (BA9)",
                                     "Brain - Cortex",
                                     "Brain - Hippocampus",
                                     "Brain - Substantia nigra",
                                     "Brain - Amygdala",
                                     "Brain - Putamen (basal ganglia)",
                                     "Brain - Caudate (basal ganglia)",
                                     "Brain - Hypothalamus",
                                     "Brain - Nucleus accumbens (basal ganglia)") ) ) )

# define colors of the clusers
cols <- RColorBrewer::brewer.pal(8, "Accent")

CountClust::StructureGGplot(omega = topic_clus$omega,
                annotation= annotation,
                palette = cols,
                yaxis_label = "",
                order_sample = TRUE,
                split_line = list(split_lwd = .4,
                                  split_col = "white"),
                axis_tick = list(axis_ticks_length = .1,
                                 axis_ticks_lwd_y = .1,
                                 axis_ticks_lwd_x = .1,
                                 axis_label_size = 3,
                                 axis_label_face = "bold"))


```

We go one level up. 

```{r}
libsize_brain <- rowSums(z_level_1_counts);
omega_brain_2 <- topic_clus$omega;  theta_brain_2 <- topic_clus$theta;
lambda_brain_2 <- sweep(omega_brain_2, 1, libsize_brain, "*");

z_level_2 <- array(0,c(dim(lambda_brain_2)[1], dim(theta_brain_2)[1], dim(lambda_brain_2)[2]));
for(k in 1:dim(theta_brain_2)[2]){
  z_level_2[,,k] <- lambda_brain_2[,k]%*%t(theta_brain_2[,k])
}

z_level_2_counts <- floor(apply(z_level_2, c(1,3), sum));
```

We apply topic model with $K3=2$ on this matrix.

```{r}
topic_clus_2 <- maptpx::topics(z_level_2_counts, K=2, tol=0.1);
```

```{r}
rownames(omega_brain) <- paste0("X", 1:length(brain_labels))
annotation <- data.frame(
    sample_id = paste0("X", 1:length(brain_labels)),
    tissue_label = factor(brain_labels,
                          levels = rev(c("Brain - Cerebellar Hemisphere",
                                     "Brain - Cerebellum",
                                     "Brain - Spinal cord (cervical c-1)",
                                     "Brain - Anterior cingulate cortex (BA24)",
                                     "Brain - Frontal Cortex (BA9)",
                                     "Brain - Cortex",
                                     "Brain - Hippocampus",
                                     "Brain - Substantia nigra",
                                     "Brain - Amygdala",
                                     "Brain - Putamen (basal ganglia)",
                                     "Brain - Caudate (basal ganglia)",
                                     "Brain - Hypothalamus",
                                     "Brain - Nucleus accumbens (basal ganglia)") ) ) )

# define colors of the clusers
cols <- RColorBrewer::brewer.pal(8, "Accent")

CountClust::StructureGGplot(omega = topic_clus_2$omega,
                annotation= annotation,
                palette = cols,
                yaxis_label = "",
                order_sample = TRUE,
                split_line = list(split_lwd = .4,
                                  split_col = "white"),
                axis_tick = list(axis_ticks_length = .1,
                                 axis_ticks_lwd_y = .1,
                                 axis_ticks_lwd_x = .1,
                                 axis_label_size = 3,
                                 axis_label_face = "bold"))


```


What happens if we apply admixture separately for $K=2$ and $K=3$ on the data?

```{r echo=TRUE, eval=FALSE}
topic_clus_brain <- maptpx::topics(t(brain_data), K=2, tol=100);
save(topic_clus_brain, file="../rdas/topics_brain_gtex_v6.k.2.rda")
```

```{r}
topic_clus_brain <- get(load("../rdas/topics_brain_gtex_v6.k.2.rda"));

omega <- topic_clus_brain$omega;

# define colors of the clusers
cols <- RColorBrewer::brewer.pal(8, "Accent")

CountClust::StructureGGplot(omega = omega,
                annotation= annotation,
                palette = cols,
                yaxis_label = "",
                order_sample = TRUE,
                split_line = list(split_lwd = .4,
                                  split_col = "white"),
                axis_tick = list(axis_ticks_length = .1,
                                 axis_ticks_lwd_y = .1,
                                 axis_ticks_lwd_x = .1,
                                 axis_label_size = 3,
                                 axis_label_face = "bold"))


```

## GTEx whole data analysis

```{r echo=TRUE, eval=TRUE}
gtex_17 <- get(load("../rdas/gtexv6fit.k.17.rda"));
gtex_17_omega <- gtex_17$omega
gtex_17_theta <- gtex_17$theta
```

We perform Structure plot visualization

```{r}
omega <- gtex_17_omega
colnames(omega) <- c(1:NCOL(omega))

# make cell sample labels
# want a version consistent with majority of the literature
sample_labels <- read.table("../external_data/GTEX_V6/samples_id.txt",
                            header = TRUE, sep = " ",
                            stringsAsFactors = FALSE)
tissue_labels <- vector("numeric", NROW(sample_labels))
tissue_labels <- sample_labels[ ,3]

# clean labels
tissue_labels[grep("Nucleus", tissue_labels)] <- "Brain -N. accumbens"
tissue_labels[grep("Putamen", tissue_labels)] <- "Brain -Putamen"
tissue_labels[grep("Caudate", tissue_labels)] <- "Brain -Caudate"
tissue_labels[grep("Gastroe", tissue_labels)] <- "Esophagus -Gastroesophageal Jn."
tissue_labels[grep("cingulate", tissue_labels)] <- "Brain - Anterior cortex (BA24)."
tissue_labels[grep("EBV", tissue_labels)] <- "Cells -EBV-lymphocytes"
tissue_labels[grep("Suprapubic", tissue_labels)] <- "Skin - Unexposed (Suprapubic)"
tissue_labels[grep("Lower Leg", tissue_labels)] <- "Skin - Sun Exposed (Lower Leg)"

# find sample orders in hierarchical clustering
docweights_per_tissue_mean <- apply(omega, 2,
                                    function(x) { tapply(x, tissue_labels, mean) })
ordering <- heatmap(docweights_per_tissue_mean)$rowInd

# order tissue by hierarhical clustering results
tissue_levels_reordered <- unique(tissue_labels)[ordering]


annotation <- data.frame(
    sample_id = paste0("X", 1:length(tissue_labels)),
    tissue_label = factor(tissue_labels,
                          levels = rev(tissue_levels_reordered ) ) )


cols1 <- c(rev(RColorBrewer::brewer.pal(12, "Paired"))[c(3,4,7,8,11,12,5,6,9,10)],
           RColorBrewer::brewer.pal(12, "Set3"))

CountClust::StructureGGplot(omega = omega,
                annotation= annotation,
                palette = cols1,
                yaxis_label = "",
                order_sample = TRUE,
                split_line = list(split_lwd = .1,
                                  split_col = "white"),
                axis_tick = list(axis_ticks_length = .1,
                                 axis_ticks_lwd_y = .1,
                                 axis_ticks_lwd_x = .1,
                                 axis_label_size = 5,
                                 axis_label_face="bold"))

```

### metatopic for K=12

```{r echo=TRUE, eval=FALSE}
libsize_gtex <- colSums(matdata);

omega_1 <- gtex_17_omega
theta_1 <- gtex_17_theta

lambda_1 <- sweep(omega_1, 1, libsize_gtex, "*");

z_level_1 <- array(0,c(dim(lambda_1)[1], dim(theta_1)[1], dim(lambda_1)[2]));
for(k in 1:dim(theta_1)[2]){
  z_level_1[,,k] <- lambda_1[,k]%*%t(theta_1[,k])
}

z_level_1_counts <- apply(z_level_1, c(1,3), function(x) return(floor(sum(x))));

save(z_level_1_counts, file="../rdas/gtex_level_2_data.rda")
```

We apply topic model with $K2=3$ on this matrix.

```{r}
z_level_1_counts <- get(load("../rdas/gtex_level_2_data.rda"))
topic_clus <- maptpx::topics(z_level_1_counts, K=12, tol=0.1);
```

```{r}
omega <- topic_clus$omega
# define colors of the clusers
cols1 <- c(rev(RColorBrewer::brewer.pal(12, "Paired"))[c(3,4,7,8,11,12,5,6,9,10)],
           RColorBrewer::brewer.pal(12, "Set3"))

CountClust::StructureGGplot(omega = omega,
                annotation= annotation,
                palette = cols1,
                yaxis_label = "",
                order_sample = TRUE,
                split_line = list(split_lwd = .4,
                                  split_col = "white"),
                axis_tick = list(axis_ticks_length = .1,
                                 axis_ticks_lwd_y = .1,
                                 axis_ticks_lwd_x = .1,
                                 axis_label_size = 3,
                                 axis_label_face = "bold"))


```

### metatopic K=10

```{r echo=TRUE, eval=FALSE}
libsize_gtex <- rowSums(z_level_1_counts);

omega_2 <- topic_clus$omega
theta_2 <- topic_clus$theta

lambda_2 <- sweep(omega_2, 1, libsize_gtex, "*");

z_level_2 <- array(0,c(dim(lambda_2)[1], dim(theta_2)[1], dim(lambda_2)[2]));
for(k in 1:dim(theta_2)[2]){
  z_level_2[,,k] <- lambda_2[,k]%*%t(theta_2[,k])
}

z_level_2_counts <- apply(z_level_2, c(1,3), function(x) return(floor(sum(x))));

```

We apply topic model with $K2=3$ on this matrix.

```{r}
topic_clus <- maptpx::topics(z_level_2_counts, K=10, tol=0.1);
```

```{r}
omega <- topic_clus$omega
# define colors of the clusers
cols1 <- c(rev(RColorBrewer::brewer.pal(12, "Paired"))[c(3,4,7,8,11,12,5,6,9,10)],
           RColorBrewer::brewer.pal(12, "Set3"))

CountClust::StructureGGplot(omega = omega,
                annotation= annotation,
                palette = cols1,
                yaxis_label = "",
                order_sample = TRUE,
                split_line = list(split_lwd = .4,
                                  split_col = "white"),
                axis_tick = list(axis_ticks_length = .1,
                                 axis_ticks_lwd_y = .1,
                                 axis_ticks_lwd_x = .1,
                                 axis_label_size = 3,
                                 axis_label_face = "bold"))


```


### metatopic K=5

```{r echo=TRUE, eval=FALSE}
libsize_gtex <- rowSums(z_level_2_counts);

omega_3 <- topic_clus$omega
theta_3 <- topic_clus$theta

lambda_3 <- sweep(omega_3, 1, libsize_gtex, "*");

z_level_3 <- array(0,c(dim(lambda_3)[1], dim(theta_3)[1], dim(lambda_3)[2]));
for(k in 1:dim(theta_3)[2]){
  z_level_3[,,k] <- lambda_3[,k]%*%t(theta_3[,k])
}

z_level_3_counts <- apply(z_level_3, c(1,3), function(x) return(floor(sum(x))));

```

We apply topic model with $K2=3$ on this matrix.

```{r}
topic_clus <- maptpx::topics(z_level_3_counts, K=6, tol=0.1);
```

```{r}
omega <- topic_clus$omega
# define colors of the clusers
cols1 <- c(rev(RColorBrewer::brewer.pal(12, "Paired"))[c(3,4,7,8,11,12,5,6,9,10)],
           RColorBrewer::brewer.pal(12, "Set3"))

CountClust::StructureGGplot(omega = omega,
                annotation= annotation,
                palette = cols1,
                yaxis_label = "",
                order_sample = TRUE,
                split_line = list(split_lwd = .4,
                                  split_col = "white"),
                axis_tick = list(axis_ticks_length = .1,
                                 axis_ticks_lwd_y = .1,
                                 axis_ticks_lwd_x = .1,
                                 axis_label_size = 3,
                                 axis_label_face = "bold"))


```


### metatopic K=3

```{r echo=TRUE, eval=FALSE}
libsize_gtex <- rowSums(z_level_3_counts);

omega_4 <- topic_clus$omega
theta_4 <- topic_clus$theta

lambda_4 <- sweep(omega_4, 1, libsize_gtex, "*");

z_level_4 <- array(0,c(dim(lambda_4)[1], dim(theta_4)[1], dim(lambda_4)[2]));
for(k in 1:dim(theta_4)[2]){
  z_level_4[,,k] <- lambda_4[,k]%*%t(theta_4[,k])
}

z_level_4_counts <- apply(z_level_4, c(1,3), function(x) return(floor(sum(x))));

```

We apply topic model with $K2=3$ on this matrix.

```{r}
topic_clus <- maptpx::topics(z_level_4_counts, K=3, tol=0.1);
```

```{r}
omega <- topic_clus$omega
# define colors of the clusers
cols1 <- c(rev(RColorBrewer::brewer.pal(12, "Paired"))[c(3,4,7,8,11,12,5,6,9,10)],
           RColorBrewer::brewer.pal(12, "Set3"))

CountClust::StructureGGplot(omega = omega,
                annotation= annotation,
                palette = cols1,
                yaxis_label = "",
                order_sample = TRUE,
                split_line = list(split_lwd = .4,
                                  split_col = "white"),
                axis_tick = list(axis_ticks_length = .1,
                                 axis_ticks_lwd_y = .1,
                                 axis_ticks_lwd_x = .1,
                                 axis_label_size = 3,
                                 axis_label_face = "bold"))


```


### metatopic K=2

```{r echo=TRUE, eval=FALSE}
libsize_gtex <- rowSums(z_level_4_counts);

omega_5 <- topic_clus$omega
theta_5 <- topic_clus$theta

lambda_5 <- sweep(omega_5, 1, libsize_gtex, "*");

z_level_5 <- array(0,c(dim(lambda_5)[1], dim(theta_5)[1], dim(lambda_5)[2]));
for(k in 1:dim(theta_5)[2]){
  z_level_5[,,k] <- lambda_5[,k]%*%t(theta_5[,k])
}

z_level_5_counts <- apply(z_level_5, c(1,3), function(x) return(floor(sum(x))));

```

We apply topic model with $K2=2$ on this matrix.

```{r}
topic_clus <- maptpx::topics(z_level_5_counts, K=2, tol=0.1);
```

```{r}
omega <- topic_clus$omega
# define colors of the clusers
cols1 <- c(rev(RColorBrewer::brewer.pal(12, "Paired"))[c(3,4,7,8,11,12,5,6,9,10)],
           RColorBrewer::brewer.pal(12, "Set3"))

CountClust::StructureGGplot(omega = omega,
                annotation= annotation,
                palette = cols1,
                yaxis_label = "",
                order_sample = TRUE,
                split_line = list(split_lwd = .4,
                                  split_col = "white"),
                axis_tick = list(axis_ticks_length = .1,
                                 axis_ticks_lwd_y = .1,
                                 axis_ticks_lwd_x = .1,
                                 axis_label_size = 3,
                                 axis_label_face = "bold"))


```
