---
title: "GTEX V6 analysis"
author: "Kushal K Dey"
date: "January 29, 2016"
output: 
  html_document:
    css: floating-toc.css
    toc: true
---

## Objective

In this script, we perform the topic model and Structure plots on the GTEX V6 data across all tissues and then exclusively on the brain tissues. We fit a topic model with $K=15$ topics over all tissues in the GTEX study and then plot the Structure plot. We also fit a topic model with $4$ topics on the brain data exclusively and then plot the corresponding Structure plot as well. 

```{r echo=TRUE, eval=TRUE, warning=FALSE, message=FALSE}
rm(list=ls())
library(data.table)
#install_github('kkdey/maptpx') 
library(maptpx)
library(CountClust)
library(data.table)

```

## Data preprocessing 

```{r echo=TRUE, eval=TRUE, warning=FALSE}
data <- data.frame(fread('../external_data/GTEX_V6/cis_gene_expression.txt'));
matdata <- data[,-(1:2)];
samples_id=read.table("../external_data/GTEX_V6/samples_id.txt")[,3];
```

We fit a topic model for $K=15$ over all the tissues in the GTEX V6 data.

```{r echo=TRUE, eval=FALSE}
Topic_Clus=topics(t(matdata),15,kill=0,tol=0.1);
docweights=Topic_Clus$omega;
```

## Structure Plot (All tissues)

```{r echo=TRUE, eval=TRUE}
docweights <- as.matrix(read.table("../external_data/GTEX_V6/admix_out_GTEX_V6/omega_cis_genes_0_1_2.txt"));
K=dim(docweights)[2];
color=c("red","blue","cornflowerblue","black","cyan","darkblue",
        "brown4","burlywood","darkgoldenrod1","darkgray","deepskyblue","darkkhaki",
        "firebrick","darkorchid","hotpink","green","magenta","yellow", "azure1","azure4");

samples_id <- as.character(samples_id)
samples_id[grep("Nucleus", samples_id)] = "Brain -N. accumbens (basal ganglia)"
samples_id[grep("Gastroe", samples_id)] = "Esophagus -Gastroesophageal Jn."
samples_id[grep("cingulate", samples_id)] = "Brain - Anterior cortex (BA24)."
#samples_id[grep("Brain", samples_id)] = "Brain"
samples_id <- as.factor(samples_id)

docweights_per_tissue_mean <- apply(docweights,2,function(x) tapply(x,samples_id,mean));
ordering=heatmap(docweights_per_tissue_mean)$rowInd;

unique_samples_id_ordered = unique(samples_id)[ordering];

clus_ordered =unlist(lapply(1:53, function(x) which(samples_id == unique_samples_id_ordered[x])));


docweights_ordering = docweights[clus_ordered,];
samples_id_ordered = samples_id[clus_ordered];
indices1 <- which(samples_id_ordered =="Artery - Coronary");
indices2 <- 1:7667;
indices3 <- 7668:8227;
indices4 <- 8361:8555;

indices <- c(indices2, indices1, indices3, indices4);
samples_id_ordered <- samples_id_ordered[indices];
samples_id_ordered_all <- samples_id_ordered;
docweights_ordering <- docweights_ordering[indices,]

#png(paste0('../plots/fig1_paper/temp1.png'),width=700,height=300)
par(mar=c(15,3,2,1))
barplot(t(docweights_ordering),col=color[1:K],width=1, axisnames=F,space=0,border=NA,main=paste("Unthinned data: K=",K),las=1,ylim=c(0,1),cex.axis=0.9,cex.main=1)

labels = match(unique(samples_id_ordered), samples_id_ordered);
abline(v=labels)

labels_low=labels;
labels_up=c(labels[2:length(labels)],dim(docweights_ordering)[1]);
mid_point=labels_low +0.5*(labels_up-labels_low);

axis(1,at=mid_point, unique(samples_id_ordered),las=3, cex.axis=0.8);
#dev.off()

```

## Filtering Brain data 

```{r echo=TRUE, eval=TRUE}
brain_indices <- grep("Brain", samples_id);

brain_data <- matdata[,brain_indices];
colnames(brain_data) <- samples_id[brain_indices];
brain_data_frame <- cbind.data.frame(data[,2],brain_data);
```

## Topic model fit (Brain Data)

We apply the topic model on the brain data.

```{r echo=TRUE, eval=FALSE}
Topic_Clus=topics(t(brain_data),4,kill=0,tol=0.005);
docweights=Topic_Clus$omega;
```

## Structure Plot (Brain Data)

```{r echo=TRUE, eval=TRUE}

docweights <- as.matrix(read.table('../external_data/GTEX_V6/admix_out_GTEX_V6/omega_cis_genes_brain.txt'));
K=dim(docweights)[2];
color=c("red","blue","cornflowerblue","black","cyan","darkblue",
        "brown4","burlywood","darkgoldenrod1","darkgray","deepskyblue","darkkhaki",
        "firebrick","darkorchid","hotpink","green","magenta","yellow", "azure1","azure4");
brain_ids <- samples_id[brain_indices];
ordering <- order(brain_ids);
samples_id_ordered <- brain_ids[ordering];
docweights_ordering <- docweights[ordering,];
#png(filename=paste0('../plots/GTEX_V6_brain_thin_',0,'.png'),width=700,height=300)
par(mar=c(14,2,2,1))
barplot(t(docweights_ordering),col=2:(K+1),axisnames=F,space=0,border=NA,main=paste("No. of clusters=",K),las=1,ylim=c(0,1),cex.axis=1,cex.main=1)

labels = match(unique(samples_id_ordered), samples_id_ordered);
abline(v=labels)

labels_low=labels;
labels_up=c(labels[2:length(labels)],dim(docweights_ordering)[1]);
mid_point=labels_low +0.5*(labels_up-labels_low);

axis(1,at=mid_point, unique(samples_id_ordered),las=2, cex.axis=0.8);
#dev.off()
```
