---
title: "identifiability check part 3"
author: "Kushal K Dey"
date: "September 1, 2015"
output: html_document
---

In this script, we try to address the identifiability problem in topic models. Is it actually not identifiable? How to deal with it?


## First set up

Let us assume we have two populations $P_1$ and $P_2$ and the corresonding $\theta$ values of relative gene expression given by 

$$\theta_1 = (0.1,0.2, 0.70/98, 0.70/98,\cdots, 0.70/98) $$
$$\theta_2 = (0.70/98, 0.70/98,\cdots, 0.70/98, 0.1,0.2) $$

We take the admixture proportions to be from $0.6$ to $0.4$ varying continuously. 

```{r echo=TRUE, eval=TRUE}

library(maptpx)
n.out <- 200
omega_sim <- cbind(seq(0.6,0.4,length.out=n.out), 1- seq(0.6,0.4,length.out=n.out));
K <- dim(omega_sim)[2];
barplot(t(omega_sim),col=2:(K+1),axisnames=F,space=0,border=NA,main=paste("No. of clusters=",K),las=1,ylim=c(0,1),cex.axis=1.5,cex.main=1.4)
```

Next we read in the $\theta$ values we reported above

```{r echo=TRUE, eval=TRUE}
freq1 <- rbind(c(0.1,0.2,rep(0.70/98,98)),c(rep(0.70/98,98), 0.1,0.2));
```

Next we generate the counts data and apply the Admixture model and plot the Structure plot.

```{r echo=TRUE, eval=TRUE}
counts <- t(do.call(cbind,lapply(1:dim(omega_sim)[1], function(x) rmultinom(1,1000,prob=omega_sim[x,]%*%freq1))));

Topic_clus <- topics(counts, K=2,tol=0.001);
K=2
barplot(t(Topic_clus$omega),col=2:(K+1),axisnames=F,space=0,border=NA,main=paste("No. of clusters=",K),las=1,ylim=c(0,1),cex.axis=1.5,cex.main=1.4)

```

Now we consider the populations of $0.6P_1 + 0.4P_2$ and $0.4P_1 + 0.6P_2$. Let $\nu_1$ and $\nu_2$ be the new gene relative expression for the two clusters. Then we can write 

$$ \nu_1 = 0.6 \theta_1 + 0.4 \theta_2 $$
$$ \nu_2 = 0.4 \theta_1 + 0.6 \theta_2 $$

We first take the admixture proportions.

```{r echo=TRUE, eval=TRUE}

library(maptpx)
n.out <- 200
omega_sim <- cbind(seq(0,1,length.out=n.out), 1- seq(0,1,length.out=n.out));
K <- dim(omega_sim)[2];
barplot(t(omega_sim),col=2:(K+1),axisnames=F,space=0,border=NA,main=paste("No. of clusters=",K),las=1,ylim=c(0,1),cex.axis=1.5,cex.main=1.4)
```

Next we define the relative gene expression.

```{r echo=TRUE, eval=TRUE}
freq2 <- matrix(0,2,length(freq1[1,]))
freq2[1,] <- 0.6*freq1[1,]+0.4*freq1[2,];
freq2[2,] <- 0.6*freq1[2,]+0.4*freq1[1,];
```


Next we generate the counts data and apply the Admixture model and plot the Structure plot.

```{r echo=TRUE, eval=TRUE}
counts <- t(do.call(cbind,lapply(1:dim(omega_sim)[1], function(x) rmultinom(1,1000,prob=omega_sim[x,]%*%freq2))));

Topic_clus <- topics(counts, K=2,tol=0.001);
K=2
barplot(t(Topic_clus$omega),col=2:(K+1),axisnames=F,space=0,border=NA,main=paste("No. of clusters=",K),las=1,ylim=c(0,1),cex.axis=1.5,cex.main=1.4)

```

The gene annotations reveal the top genes as 

```{r echo=FALSE, eval=TRUE}
topics_theta <- Topic_clus$theta;
K= dim(Topic_clus$omega)[2];

KL_score_poisson <- lapply(1:K, function(n) 
                                {
                                    out <- t(apply(topics_theta, 1, function(x)
                                                                    {
                                                                      y=x[n] *log(x[n]/x) + x - x[n];
                                                                      return(y)
                                                                     }));
                                    return(out)
                                })
                   

indices_mat_poisson=matrix(0,dim(topics_theta)[2],10);

for(k in 1:dim(topics_theta)[2])
{
  temp_mat <- KL_score_poisson[[k]][,-k];
  vec <- order(temp_mat, decreasing=TRUE)[1:4]
}

print(vec)
```

## Second set up

In this set up, we keep the gene frequencies as before but we do not vary the admixture proportions continuously, rather we keep them fixed at a particular value. 

```{r echo=TRUE, eval=TRUE}

library(maptpx)
n.out <- 200
omega_sim <- cbind(rep(0.6,n.out), 1- rep(0.6,n.out));
K <- dim(omega_sim)[2];
barplot(t(omega_sim),col=2:(K+1),axisnames=F,space=0,border=NA,main=paste("No. of clusters=",K),las=1,ylim=c(0,1),cex.axis=1.5,cex.main=1.4)
```

Next we read in the $\theta$ values we reported above

```{r echo=TRUE, eval=TRUE}
freq1 <- rbind(c(0.1,0.2,rep(0.70/98,98)),c(rep(0.70/98,98), 0.1,0.2));
```

Next we generate the counts data and apply the Admixture model and plot the Structure plot.

```{r echo=TRUE, eval=TRUE}
counts <- t(do.call(cbind,lapply(1:dim(omega_sim)[1], function(x) rmultinom(1,1000,prob=omega_sim[x,]%*%freq1))));

Topic_clus <- topics(counts, K=2,tol=0.001);
K=2
barplot(t(Topic_clus$omega),col=2:(K+1),axisnames=F,space=0,border=NA,main=paste("No. of clusters=",K),las=1,ylim=c(0,1),cex.axis=1.5,cex.main=1.4)

```


Next we consider the analog of the second example of set up 1. Here the admixture proportion is taken to be all 1 for the first cluster. 

```{r echo=TRUE, eval=TRUE}

library(maptpx)
n.out <- 200
omega_sim <- cbind(rep(1,n.out), 1- rep(1,n.out));
K <- dim(omega_sim)[2];
barplot(t(omega_sim),col=2:(K+1),axisnames=F,space=0,border=NA,main=paste("No. of clusters=",K),las=1,ylim=c(0,1),cex.axis=1.5,cex.main=1.4)
```


However the relative gene expression is of the form $0.6P_1 + 0.4P_2$ and $0.4P_1 + 0.6P_2$.

```{r echo=TRUE, eval=TRUE}
freq2 <- matrix(0,2,length(freq1[1,]))
freq2[1,] <- 0.6*freq1[1,]+0.4*freq1[2,];
freq2[2,] <- 0.6*freq1[2,]+0.4*freq1[1,];
```

Next we generate the counts data and apply the Admixture model and plot the Structure plot.

```{r echo=TRUE, eval=TRUE}
counts <- t(do.call(cbind,lapply(1:dim(omega_sim)[1], function(x) rmultinom(1,1000,prob=omega_sim[x,]%*%freq2))));

Topic_clus <- topics(counts, K=2,tol=0.001);
K=2
barplot(t(Topic_clus$omega),col=2:(K+1),axisnames=F,space=0,border=NA,main=paste("No. of clusters=",K),las=1,ylim=c(0,1),cex.axis=1.5,cex.main=1.4)

```

The gene annotations reveal the top genes as 

```{r echo=FALSE, eval=TRUE}
topics_theta <- Topic_clus$theta;
K= dim(Topic_clus$omega)[2];

KL_score_poisson <- lapply(1:K, function(n) 
                                {
                                    out <- t(apply(topics_theta, 1, function(x)
                                                                    {
                                                                      y=x[n] *log(x[n]/x) + x - x[n];
                                                                      return(y)
                                                                     }));
                                    return(out)
                                })
                   

indices_mat_poisson=matrix(0,dim(topics_theta)[2],10);

for(k in 1:dim(topics_theta)[2])
{
  temp_mat <- KL_score_poisson[[k]][,-k];
  vec <- order(temp_mat, decreasing=TRUE)[1:4]
}

print(vec)
```
