---
title: 'Identifiability check: Part 2'
author: "Kushal K Dey"
date: "August 7, 2015"
output: html_document
---

## Topic model with mixture prop from 0 to 1 

```{r}

library(maptpx)
n.out <- 200
omega_sim <- cbind(seq(1,0,length.out=n.out), 1- seq(1,0,length.out=n.out));
K <- dim(omega_sim)[2];
png(filename = '../plots/true_structure_setup_1.png')
barplot(t(omega_sim),col=2:(K+1),axisnames=F,space=0,border=NA,main=paste("True structure plot, K=",K),las=1,ylim=c(0,1),cex.axis=1.5,cex.main=1.4)
dev.off()
```

```{r, echo=TRUE, eval=TRUE}
freq <- rbind(c(0.01,0.05,rep(0.95/498,498)),c(0.01,rep(0.95/498,498),0.05));
```

```{r, echo=TRUE, eval=TRUE}
counts <- t(do.call(cbind,lapply(1:dim(omega_sim)[1], function(x) rmultinom(1,1000,prob=omega_sim[x,]%*%freq))));
```

```{r, echo=TRUE, eval=TRUE}
Topic_clus <- topics(counts, K=2,tol=0.001);
K=2
png(filename = '../plots/est_structure_setup_1.png')
barplot(t(Topic_clus$omega),col=2:(K+1),axisnames=F,space=0,border=NA,main=paste("Est. structure plot, K=",K),las=1,ylim=c(0,1),cex.axis=1.5,cex.main=1.4)
dev.off()

```

Which genes are highly expressed for the last topic model? are they the same genes? 

```{r gene_annotations_1, echo=TRUE, eval=TRUE}
topics_theta <- Topic_clus$theta;
K= dim(Topic_clus$omega)[2];

KL_score_poisson <- lapply(1:K, function(n) 
                                {
                                    out <- t(apply(topics_theta, 1, function(x)
                                                                    {
                                                                      y=x[n] *log(x[n]/x) + x - x[n];
                                                                      return(y)
                                                                     }));
                                    return(out)
                                })
                   

indices_mat_poisson=matrix(0,dim(topics_theta)[2],10);

for(k in 1:dim(topics_theta)[2])
{
  temp_mat <- KL_score_poisson[[k]][,-k];
  vec <- order(temp_mat, decreasing=TRUE)[1:4]
  print(vec)
}

## so under this scenario, the genes that is shown by Structure are also not reliable: low signal?

```

## Topic model with mixture prop from 0.4 to 0.6 (same freq as above)

```{r}

library(maptpx)
n.out <- 200
omega_sim <- cbind(seq(0.6,0.4,length.out=n.out), 1- seq(0.6,0.4,length.out=n.out));
K <- dim(omega_sim)[2];
png(filename = '../plots/true_structure_setup_2.png')
barplot(t(omega_sim),col=2:(K+1),axisnames=F,space=0,border=NA,main=paste("True structure plot, K=",K),las=1,ylim=c(0,1),cex.axis=1.5,cex.main=1.4)
dev.off()

```

```{r, echo=TRUE, eval=TRUE}
freq <- rbind(c(0.01,0.05,rep(0.94/498,498)),c(0.05,0.01,rep(0.94/498,498)));
```

```{r, echo=TRUE, eval=TRUE}
counts <- t(do.call(cbind,lapply(1:dim(omega_sim)[1], function(x) rmultinom(1,1000,prob=omega_sim[x,]%*%freq))));
```

```{r, echo=TRUE, eval=TRUE}
Topic_clus <- topics(counts, K=2,tol=0.001);
K=2
png(filename = '../plots/est_structure_setup_2.png')
barplot(t(Topic_clus$omega),col=2:(K+1),axisnames=F,space=0,border=NA,main=paste("Est. structure plot, K=",K),las=1,ylim=c(0,1),cex.axis=1.5,cex.main=1.4)
dev.off()


```

Which genes are highly expressed for the last topic model? are they the same genes? 

```{r gene_annotations_2, echo=TRUE, eval=TRUE}
topics_theta <- Topic_clus$theta;
K= dim(Topic_clus$omega)[2];

KL_score_poisson <- lapply(1:K, function(n) 
                                {
                                    out <- t(apply(topics_theta, 1, function(x)
                                                                    {
                                                                      y=x[n] *log(x[n]/x) + x - x[n];
                                                                      return(y)
                                                                     }));
                                    return(out)
                                })
                   

indices_mat_poisson=matrix(0,dim(topics_theta)[2],10);

for(k in 1:dim(topics_theta)[2])
{
  temp_mat <- KL_score_poisson[[k]][,-k];
  vec <- order(temp_mat, decreasing=TRUE)[1:4]
  print(vec)
}

## so under this scenario, the genes that is shown by Structure are also not reliable: low signal?

```

## Hierarchical clustering 

```{r}

library(maptpx)
n.out <- 50
omega_sim <- cbind(seq(1,0,length.out=n.out), 1- seq(1,0,length.out=n.out));
K <- dim(omega_sim)[2];
barplot(t(omega_sim),col=2:(K+1),axisnames=F,space=0,border=NA,main=paste("No. of clusters=",K),las=1,ylim=c(0,1),cex.axis=1.5,cex.main=1.4)
```

```{r, echo=TRUE, eval=TRUE}
freq <- rbind(c(0.01,0.05,rep(0.94/498,498)),c(0.01,rep(0.94/498,498),0.05));
```

```{r, echo=TRUE, eval=TRUE}
counts <- t(do.call(cbind,lapply(1:dim(omega_sim)[1], function(x) rmultinom(1,1000,prob=omega_sim[x,]%*%freq))));
```

The hierarchical plot heat map is given by 

```{r, echo=TRUE, eval=TRUE}
library(gplots)
library(philentropy)
col = c(rgb(seq(0,1,length=15),1,seq(0,1,length=15)), rgb(1,seq(1,0,length=15),seq(1,0,length=15)));
heatmap.2(distance(counts,method="euclidean"),scale="none", trace="none", distfun=function(x) dist(x,method="euclidean"), col=col);
```


The Structure plot heat map is given by 

```{r, echo=TRUE, eval=TRUE}
Topic_clus <- topics(counts, K=2,tol=0.00001);
barplot(t(Topic_clus$omega),col=2:(K+1),axisnames=F,space=0,border=NA,main=paste("No. of clusters=",K),las=1,ylim=c(0,1),cex.axis=1.5,cex.main=1.4)
heatmap.2(distance(Topic_clus$omega,method="euclidean"),scale="none", trace="none", distfun=function(x) dist(x,method="euclidean"), col=col);
```
