---
title: "Validation of kkdey/maptpx wrt Taddylab/maptpx"
author: "Kushal K Dey"
date: "April 6, 2016"
output: html_document
---

## Overview

I made some changes in the tpxEM() functionality and additionally some minor changes to the tpxlpost() function in Matt's package in order to solve the long vectors problem in FORTRAN and C. I want to validate the method on data examples on the previous version of Matt's package (which now corresponds to admix=2) with respect to my modified version (which corresponds to admix=1 and is now the default).

## Simulation Design

Assume K = 2 hidden subpopulations in the data, and n = 600 cells. We set a 600 by 2 membership weight matrix $\mathbf{q} with $\sum_{K=2} q_{n.} = 1$, where $q_{n.}$ denotes the vector of membership weight (grade?) of cell $n$. 


```{r echo=TRUE, eval=TRUE}
n.out <- 400
omega_sim <- rbind( cbind( rep(1, n.out), rep(0, n.out)), 
                    cbind( rep(0, n.out), rep(1, n.out)),
                    cbind( seq(0.6, 0.4, length.out = n.out), 
                           1- seq(0.6, 0.4,length.out=n.out)) )
dim(omega_sim)
```


```{r}
K <- dim(omega_sim)[2]
barplot(t(omega_sim), 
        col = 2:(K+1),
        axisnames = F, space = 0, 
        border = NA, 
        main=paste("No. of clusters=", K),
        las=1, ylim = c(0,1), cex.axis=1.5,cex.main=1.4)
```


We generate the cluster or topic distributions. These are also known as relative gene expression weights of individual cells. Here we simulate $G = 100$ genes, and membership weight

$\begin{pmatrix} q_{,11} & q_{1,2} & \dots & q_{1,G} \\
  q_{2,1} & q_{2,2} & \dots & q_{2,G} 
  \end{pmatrix}$


```{r echo=TRUE, eval=TRUE}
freq <- rbind( c(0.1, 0.2, rep(0.70/998, 998)),
               c(rep(0.70/998,998), 0.1, 0.2) )
str(freq)
```


Now we generate the counts table given the topic proportions matrix and the topic distributions.

For every cell $n$, we assume that $C_{n.}$ the vector of read counts coming from genes 1 to G follows a multinomial distribution where $\left( p_{n,1}, \dots, p_{n,G} \right)$ the probabilities of a read coming from genes 1 to $G$ sum up to 1. That is, 

$C_{n.} \sim Multinomial \left( C_{n+}, p_{n.} \right)$ 

and 

$p_{n.}= \sum_{k=1}^{K} q_{nk} \theta_{kg}$.


```{r echo=TRUE, eval=TRUE}
counts <- t( do.call(cbind,
                     lapply(1:dim(omega_sim)[1], 
                            function(x) 
                                rmultinom(1,1000,prob=omega_sim[x,]%*%freq))))
dim(counts)
```


## kkdey/maptpx application

```{r echo=TRUE, eval=TRUE}
system.time(Topic_clus <- maptpx::topics(counts, K=2, tol=0.001));
```

```{r echo=TRUE, eval=TRUE, message=FALSE, warning=FALSE}
# These are estimated gene relative expression
docweights <- Topic_clus$omega

library(permute)
library(BioPhysConnectoR)

# Decide on the correspondance of the simulated
# and of the estimated relative expression matrix
# 
# We consider all possible cluster ordering in the 
# estimated relative expression matrix and select
# the relative expression matrix with the nearest
# distance to the simulate relative experssion matrix
# (forbieus norm)

perm_set <- rbind(1:K,allPerms(1:K))
diff <- array(0,dim(perm_set)[1]) 
for (p in 1:dim(perm_set)[1])
{
    temp <- docweights[, perm_set[p,]]
    diff[p] <- fnorm(temp,omega_sim)
}

p_star <- which(diff==min(diff))
docweights <- docweights[,perm_set[p_star,]]

barplot(t(docweights),
        col = 2:(K+1),
        axisnames = F, space = 0, border = NA, 
        main=paste("No. of clusters=", K),
        las=1, ylim=c(0,1), cex.axis=1.5, cex.main=1.4)
```

## kkdey/maptpx application

```{r echo=TRUE, eval=TRUE}
source("../../../maptpx/R/topics.R")
source("../../../maptpx/R/tpx.R")
source("../../../maptpx/R/count.R")
library(slam)
system.time(Topic_clus <- topics(counts, K=2, tol=0.001, method_admix=1));
```

```{r echo=TRUE, eval=TRUE, message=FALSE, warning=FALSE}
# These are estimated gene relative expression
docweights <- Topic_clus$omega

library(permute)
library(BioPhysConnectoR)

# Decide on the correspondance of the simulated
# and of the estimated relative expression matrix
# 
# We consider all possible cluster ordering in the 
# estimated relative expression matrix and select
# the relative expression matrix with the nearest
# distance to the simulate relative experssion matrix
# (forbieus norm)

perm_set <- rbind(1:K,allPerms(1:K))
diff <- array(0,dim(perm_set)[1]) 
for (p in 1:dim(perm_set)[1])
{
    temp <- docweights[, perm_set[p,]]
    diff[p] <- fnorm(temp,omega_sim)
}

p_star <- which(diff==min(diff))
docweights <- docweights[,perm_set[p_star,]]

barplot(t(docweights),
        col = 2:(K+1),
        axisnames = F, space = 0, border = NA, 
        main=paste("No. of clusters=", K),
        las=1, ylim=c(0,1), cex.axis=1.5, cex.main=1.4)
```
