\section{Results}

%An outline for results  (under consideration)
%
%\begin{itemize}
%
%\item Form two separate subsections, one for the GTEx Version 4 data and the other for the single cell Zeisel data.
%
%\item For GTEx data, give a figure comprising of 4 Structure plots for different $K$s, may be $2,5,10,15$. Fix the thinning parameter $p_{thin}$ to say $0.0001$.  Also record the log likelihoods (Bayes factors) for each of the 4 models, as reported by \textbf{maptpx}. 
%
%\item Have one figure showing the robustness of the clustering method on the thinning parameter $p_{thin}$. Fix $k=10$ and vary $p_{thin}$ to be $0.0001$, $0.001$ and $0.01$. 
%
%\item One t-SNE plot for GTEx samples (with and without admixture in the same plot). Should this be in results or in discussions? Also the t-SNE probably would require an electronic supplemental file as I would need the \textbf{qtlcharts} highlighting for those plots. 
%
%\item The GTEX brain samples Structure plot  for $K=4$ that shows the neuron cell types in brain cerebellum and cerebellar hemisphere. That is to show that the clusters are driven by cell types.
%
%\item Gene annotations for the GTEx significant genes (for brain) and also for the general set up (to decide on which $K$ to fix). Use Bayes Factor?
%
%\item The Structure plot for Zeisel single cell data. again Multiple $K=2,5,7,10$. 
%
%\item Gene annotations for the Zeisel single cell data. Need to choose the optimal $K$. Use Bayes Factor?
%
%\item t-SNE plot of the admixture proportions??..Is that required? Depends on how we present t-SNE. If this goes to discussion, we will avoid it here
%
%\end{itemize}

To briefly summarize the approach, assume RNA-seq data have been summarized by a table of counts $C_{N \times G} = (c_{ng})$, where $c_{ng}$ is the number of reads from sample $n$ mapped to gene (or transcript) $g$ \cite{Oshlack2010}.  The grade of membership (GoM) model assumes that 
\begin{equation}
c_{n\cdot} \sim \text{Mult}(c_{n+}, p_{n\cdot})
\end{equation}
where $c_{n\cdot}$ denotes the count vector for the $n$th sample, $c_{n+} := \sum_g c_{ng}$, and $p_{n\cdot}$ is a probability vector (non-negative entries summing to 1) whose $g$th element represents the relative expression of gene $g$ in sample $n$. 
The model further assumes that 
\begin{equation}
p_{ng} = \sum_{k=1}^{K} q_{nk}\theta_{kg}    
\end{equation}
where $q_{n\cdot}$ is a probability vector whose $k$th element represents the grade of membership (or ``membership proportion") of
sample $n$ in cluster $k$, and $\theta_{k\cdot}$ is a probability vector whose $g$th element represents
the relative expression of gene $g$ in cluster $k$. The number of clusters $K$ is set by the analyst, and it can be helpful to explore multiple
values of $K$ (see Discussion).

Fitting this model (see Methods) results in estimated membership proportions $q$ for each sample, and estimated expression values $\theta$ for each cluster.
We visualize the membership proportions for each sample using a ``Structure plot" \cite{Rosenberg2002}, 
which is named for its widespread use in visualizing the
results of the {\it Structure} software \cite{Pritchard2000} in population genetics.
The Structure plot represents the estimated membership proportions of each sample 
as a stacked barchart, with bars of different colors representing  different clusters. Consequently samples that have similar membership proportions have
similar amounts of each color. See Figure \ref{fig:fig1} for example.
%In addition, to help biologically interpret each cluster, we developed a method to identify which genes are most distinctively differentially expressed in each cluster (see Methods).

\subsection{Clustering human tissue samples using bulk RNA expression}

We begin by illustrating the GoM model on bulk RNA expression measurements from the GTEx project (V6 dbGaP accession phs000424.v6.p1, release date: Oct 19, 2015, \url{http://www.gtexportal.org/home/}).  These data consist of per-gene read counts from RNA-seq performed on $8,555$ samples collected from $450$ human donors across $51$ tissues, lymphoblastoid cell lines, and transformed fibroblast cell-lines. We analyzed $16,069$ genes that satisfied filters (e.g. exceeding certain minimum expression levels) that were used during eQTL analyses by the GTEx project (gene list available in \url{https://github.com/stephenslab/count-clustering/blob/master/project/utilities/gene_names_GTEX_V6.txt}). 

To assess structure in these data we applied the GoM model with $K=10,12,15$. Although results differ with $K$, many of the primary patterns were consistent across $K$. Here, for brevity, we focus on results for $K=15$, shown as a Structure plot in \textbf{Figure \ref{fig:fig1}(a)} (see also an alternative visualization using a 2-dimensional projection with t-SNE \cite{Maaten2008}, \cite{Maaten2014}, in \textbf{Supplementary Figure 1 \url{http://stephenslab.github.io/count-clustering/project/src/tissues_tSNE.html}}). Reassuringly, much of the structure highlighted by these results follows the known division of samples into tissues: that is, samples from the same tissue tend to have similar grades of membership across clusters. Some tissues are represented by essentially a single cluster (e.g.~Pancreas, Whole Blood), whereas other tissues are represented as a mixture of multiple clusters (e.g.~Thyroid). Furthermore, the results highlight biological similarity among some tissues by assigning similar membership proportions to samples from those tissues  For example, samples from different parts of the brain have similar memberships, as do the arteries (aorta, tibial and coronary) and skin (sun-exposed and un-exposed). 
 Samples from the tibial nerve have small but consistent amounts of membership in common with brain tissues, as well as larger amounts in common with the adipose tissues. Indeed, many tissues show membership in cluster 1 (purple) or  ``Adipose" cluster, possibly reflecting, at least in some cases, contamination with adipose cells.

 
Each cluster is characterized by a vector that contains the mean expression level for each gene. To help biologically interpret each cluster we annotate it by identifying the genes whose expression levels most strongly distinguish that cluster from the others (see Methods).  Table ~\ref{tab:tab1} summarizes the enrichment analysis of the top driving genes of each cluster for the GTEx analysis in Figure \ref{fig:fig1}a.  We also present the summary function of the top three genes for each cluster in Supplementary Table 1. Reassuringly, many results align with known biology. For example,  the light red cluster (cluster 4), which distinguishes Testis from other tissues, is enriched with  
genes responsible for sexual reproduction and spermatogenesis etc, (e.g.: \textit{PRM2}, \textit{PRM1}, \textit{PHF7}). Similarly the light green cluster (cluster 10), which primarily distinguishes Whole Blood from other tissues, is enriched with genes responsible for immune responses (e.g. \textit{CSF3R}, \textit{IFITM2}) and hemoglobin complex and oxygen transport (e.g. \textit{HBB}, \textit{HBA1}, \textit{HBA2}). Furthermore, digestion and proteolysis related genes characterize the Pancreas cluster (cluster 13, light violet), Keratin-related genes characterize the skin cluster (cluster 6, sky blue), Myosin-related genes characterize the muscle skeletal cluster (cluster 9, green), etc. In cases where a cluster occurs in multiple tissues these annotations may be particularly helpful for understanding what may be driving this co-membership. For example, the top four genes in the  light orange cluster (cluster 8), which is common to Lung, Spleen and Small Intestine - Terminal Ileum,  all code for surfactant proteins (specifically, B, A2, A1 and C).


%A field of very active interest in recent times is to estimate the proportion of different cell types in different tissues. Marker based approaches are usually adopted to validate for different cell types and get a sense of the abundance of different cell types in the tissue samples \cite{Grun2015} \cite{Palmer2005}. The admixture model is a marker free method to obtain clusters driven by cell types. 

Although global analysis of all tissues is useful for highlighting major structure in the data, it may miss finer-scale structure within tissues or among similar tissues. For example, here the global analysis allocated similar cluster memberships to all brain tissues,  and we suspected that these tissues may exhibit substructure that could be uncovered by analyzing the brain samples separately.  \textbf{Figure \ref{fig:fig1}(b)} shows the Structure plot for $K=4$ on only the Brain samples. The results highlight much finer-scale structure compared with the global analysis. Brain Cerebellum and Cerebellar hemisphere are essentially assigned to a separate cluster, which is enriched with genes related to cell periphery and communication (e.g. \textit{PKD1}, \textit{CBLN3})as well genes expressed largely in neuronal cells and playing a role in neuron differentiation (e.g. \textit{CHGB}). The spinal cord samples also show consistently strong membership in a single cluster, the top defining gene for the cluster being \textit{MBP} which is involved in myelination of nerves in the nervous system\cite{Hu2016}.  The other driving gene \textit{GFAP} takes part in system development by acting as a marker to distinguish astrocytes during development. 


%This cluster seems likely to reflect the expected high concentration of Cerebellar granule cells in cerebellar samples.
%[POssibly GRANULE CELLS?]
% (Genetic mutations in both MYH11 and ACTA2 can cause Thoracic aortic aneurysms/dissection \cite{Renard2013}.) 

The remaining samples all show membership in multiple clusters, with cortex samples being distinguished from other samples by stronger membership in a cluster (cluster3, turquoise in Figure \ref{fig:fig1}(b)) whose distinctive genes include \textit{ENC1}, which  interacts with actin and contributes to the organisation of the cytoskeleton during the specification of neural fate \cite{Hernandez1997}.



% It seems the red cluster is mainly prominent in what are called sub-cortical regions of the brain (which are immediately below the cortex).  In that sense the four clusters, although not fully representative, do seem to be driven by the cerebellum, cortex, sub-cortex and spinal cord. These are anatomically meaningful and also functionally.  Functionally, sub-cortical regions of the brain are associated with lower order thinking tasks and these are regions that are present in reptiles and birds as well. Cortex is mainly present in mammals and is used for higher order thinking tasks. Cerebellum main function is in motion learning. Also amygdala and hippocampus are anatomically very close to each other and constitute the limbic region of the brain \url{https://s3.amazonaws.com/classconnection/118/flashcards/3700118/png/screen_shot_2012-09-17_at_75637_am1347883157040-1517F5218E616F44DE2.png}. Even substantia nigra is pretty close to both of them. Also cerebellar granule cells do constitute bulk of the cerebellum and are the principal granule cells in human body (WIKI). Given that the genes we obtained as driving genes for the clusters seem to be associated with cerebellar granule cells, seems it is meaningful to call the green cluster to be driven by the granule cells. Myelin basic protein has been found to have higher expression in spinal cord cluster when compared to other clusters. Myelin increases nervous transmission and nervous transmission is very swift through the spinal cord. This could explain why MBP is one of the top distinguishing genes for the spinal cord cluster. 




%Recent stereological approaches have shown that rat cerebellum contains $> 80 \%$ neurons (Herculano-Houzel and Lent 2005) \cite{Houzel2005}, much higher than other parts of the brain. 

\subsection{Quantitative comparison with hierarchical clustering}

The GoM model is, in many ways,  complimentary to, rather than only competing with, distance-based hierarchical clustering methods.
Nonetheless, we hypothesized that the model-based nature of the GoM approach might also provide greater accuracy in detecting substructure than distance-based methods. 
We used the GTEx data to test this hypothesis. Specifically, for each pair of tissues in the GTEx data we assessed whether or not each clustering method
correctly partitioned samples into the two tissue groups (see Methods). The GoM model was substantially more accurate in this test, succeeding
in $86 \%$ of comparisons, compared with $39 \%$ for the distance-based method; Figure \ref{fig:fig2}.


\subsection{Clustering of single-cell RNA-seq data}

Recently RNA-sequencing has become viable for single cells \cite{Tang2009}, and this technology has the promise to revolutionize understanding of intra-cellular variation in expression, and regulation more generally \cite{Trapnell2015}. Although it is traditional to describe and categorize cells in terms of distinct cell-types, 
the actual architecture of cell heterogeneity may be more complex, and in some cases perhaps better captured by the more ``continuous"  GoM model. 
In this section we aim to illustrate the potential for the GoM model to be applied to single cell data.

Single-cell RNA-seq data typically involve substantially lower effective sequencing depth compared with bulk experiments, due to the lower number of molecules available to sequence in a single cell. To check robustness of the GoM model to lower sequencing depth
 we repeated analyses above after thinning the GTEx data by a factor of $10,000$ to mimic the lower sequencing depth of a typical single cell experiment.
 %per-sample depth in the 
%single-cell data from \cite{Jaitin2014} considered below. (See  \textbf{Figure \ref{fig:figS1}} and \textbf{Figure \ref{fig:figS2}} for results with other values of $p_{thin}$.) 
For the thinned GTEx data the Structure plot for $K=15$ preserves most of the major features of the original analysis on unthinned data (Supplementary Figure 2). For the accuracy comparisons with distance-based methods, both methods suffer reduced accuracy in thinned data, but the model-based method retains its superior performance. For example, when thinning by a factor of $1,000$  the success proportion in separating tissues is  $0.10$ for hierarchical clustering and $0.32$ for GoM model.

% (for $p_{thin}=0.01$, the success proportion in separating two tissues is $0.11$  for hierarchical clustering and $0.36$ for GoM model; for $p_{thin}=0.001$, the success proportion in separating tissues is  $0.10$ for hierarchical clustering and $0.32$ for GoM model).  

 %Then we generated a set of $50$ samples randomly drawn from the pooled set of samples coming from these two tissues and then observed whether the hierarchical and the admixture were separating out samples coming from the two different tissues.

Now we apply the GoM models to two single cell RNA-seq datasets, from Jaitin \textit{et al} \cite{Jaitin2014} and Deng \textit{et al} \cite{Deng2014}.  

\subsubsection{Jaitin \text{et al}, 2014}

Jaitin \textit{et al} sequenced over $4,000$ single cells from mouse spleen. Following the original authors protocol, we also filtered out 16 genes that they found to show significant batch-specific expression. Here we analyze 1041 of these cells that were categorized as $CD11c+$ in the \textit{sorting markers} column of their data (\url{http://compgenomics.weizmann.ac.il/tanay/?page_id=519}), and which had total number of reads mapping to non-ERCC genes greater than $600$. (We believe these cells correspond roughly to the $1,040$ cells in their Figure S7.)   Our hope was that applying our method to these data would identify, and perhaps refine, the cluster structure evident in 
\cite{Jaitin2014} (their Figures 2A and 2B). However, our method yielded rather different results (Figure \ref{fig:fig3}), where most cells were assigned to have membership
in several clusters. Further, the cluster membership vectors showed systematic differences among amplification batches (which in these data is also strongly correlated with sequencing batch). For example, cells in batch 1 are characterized by strong membership in the orange cluster (cluster 5) while those in batch 4 are characterized
by strong membership in both the blue and yellow clusters (2 and 6). Some adjacent batches show similar patterns - for example batches 28 and 29 have a similar visual ``palette", as do batches 32-45. The fact that batch effects are detectable in data like these is not particularly surprising. There is a growing recognition of the importance of batch effects in high-throughput data generally \cite{Leek2010} and in single cell data specifically \cite{Hicks2015}. And indeed, dimension reduction methods such as the ones we use here can be helpful in controlling for such effects \cite{Leek2007} \cite{Stegle2012}. However, why these batch effects are not evident in Figures 2A and 2B of \cite{Jaitin2014} is unclear to us. 


%identify distinct clusters, corresponding to the clusters of B cells, NK cells, pDCs and monocytes
%focusing initially on a heterogeneous mix enriched for expression of the CD11c surface marker. 
% The aim of their study was to separate out the B cells, NK cells, pDCs and monocytes.   \textbf{Fig \ref{fig:fig3}} (\textit{top panel}) presents the Structure plot  for $K=7$ for the Jaitin \textit{et al} data with the samples arranged by their amplification batch (which was a refinement of the sequencing batch). This highlights the need for caution regarding interpreting Admixture results or any clustering results, as there is a possibility of technical effects driving the clusters instead of true biological effects. There has been a growing concern among biostatisticians today about how to deal with batch effects \cite{Leek2010} \cite{Hicks2015}. \\[1 pt]

% Deng \textit{et al} collected expression data from individual cells from zygote to blastocyst stages of mouse preimplantation development \cite{Deng2014}. Deng \textit{et al}'s analysis focussed particularly on allele-specific expression from the two contributing mouse strains (CAST/EiJ and C57BL/6J). Here we analyze the counts of the two alleles combined. Visual inspection of the Principal Components Analysis in \cite{Deng2014} suggested 6-7 clusters, so we fit the cluster model with $K=6$. 
% The results (Figure \ref{fig:fig4}) clearly highlight the structure in the different development stages starting from zygote, through early/mid/late 2 cells, 4 cells, 8 cells, 16 cells, and early/mid blastocyst to finally late blastocyst. Specifically, cells that are from the same stage show similar cluster membership proportions. Further, many of the clusters show notable trends through the stages. For example, 
% membership in the green cluster is non-existent in early stages, starts in the 4-cell stage, becomes more prominent in the 8-16 cell stages, drops substantially in the early and mid-blastocyte stages, and is essentially absent in the late blastocytes. More generally, cluster memberships for cells from adjacent stages tend to be more similar to one another than those for cells from distant stages. 

% Examining the clustering results by embryo highlights apparent embryo-level effects in the early stages (Figure \ref{fig:fig4}): that is, cells from the same embryo sometimes showed distinctive differences from other embryos. For example, the two cells from one of the 2-cell embryos (check) shows much stronger membership in the magenta cluster than other 2-cell embryos, and four cells from one of the 4-cell embryos (embryo 4) shows consistently more yellow membership than the other 4-cell embryos. 

% Finally, the results indicate a few samples that appear to be outliers - for example, a cell from a 16-cell embryo that looks like a very early stage cell (zygote or early 2-cell), and a cell from an 8-stage embryo that looks rather different from any of the others.

\subsubsection{Deng \text{et al}, 2014}

Deng \textit{et al} collected single-cell expression data of mouse preimplantation embryos from the zygote to blastocyst stage \cite{Deng2014} and analysed the trend of allele-specific expression in early embryo development. Here we analyse the combined counts of the two alleles. Visual inspection of the Principal Components Analysis in \cite{Deng2014} suggested 6-7 clusters, so we focus on the GoM model with $K=6$. 

The results (Figure \ref{fig:fig4}) clearly highlight the continuous pattern of grades of membership through early embryonic development stages. Initially at the zygote stage, the embryos are represented by the blue cluster. Continuing on to the cleavage division stages, the grades of membership change to include a mixture of blue and magenta clusters at the mid 2-cell stage, a mixture of magenta and yellow clusters at the late 2-cell stage, and then a mixture of yellow and green at the 4-cell stage. Indeed, many of the clusters showed notable trends through the stages. For example, membership in the green cluster is non-existent in the early stages, but starting from the 4-cell stage, the green becomes more prominent in the 8-16 cell stages, drops substantially in the early and mid-blastocyst stages, and is essentially absent in the late blastocyst. More generally, the GoM results highlight biological similarity between embryos from adjacent stages than embryos from distant stages during preimplantation. 

Examining the clustering results by individual embryos highlights apparent embryo-level effects in the early stages (Figure \ref{fig:fig4}). 

%For example, cells from the four embryos collected at the 16-cell stage are represented by a mixture of blue, green, yellow, and purple clusters. 

While all the cells from 16-cell stage have high memberships in the green cluster, it is noteworthy that cells coming from two of the embryos at the 16-cell stage have memberships in both purple and yellow, while the other two  have memberships only in yellow cluster. Finally, the results indicate a few samples that appear to be outliers. For example, a cell from a 16-cell embryo is represented by the blue cluster - a cluster that represents cells at the zygote and early 2-cell stage. Also, a cell from an 8-stage embryo is driven by the purple cluster - a cluster that represents cells from the blastocyst stage.

%and cleavage division (e.g., \textit{Tfgb2}, \textit{Pttg1}), 

Enrichment analysis of the top driving genes in each cluster indicates that the grades of membership align with known biology during embryonic preimplantation (Table \ref{tab:tab3}). For example, the blue cluster is enriched with genes responsible for germ cell development (e.g., \textit{Bcl2l10}, \textit{Spin1}) which are essential to the zygote and 2-cell stages. Similarly, the green cluster is enriched with cytoskeletal genes (e.g., \textit{Fbox15}) and cytoplasm genes (e.g., \textit{Tceb1}, \textit{Hsp90ab1}), all of which are essential for compaction at the 8-cell stage and morula formation at the 16-cell stage. Furthermore, the results showed some clusters formed a biologically meaningful mixture. For example, cells in the blastocyst are represented by a mixture of orange and purple clusters. The orange cluster seems to be  enriched with genes involving in the formation of outer trophoblast cells (e.g., \textit{Tspan8}, \textit{Krt8}, \textit{Id2} \cite{Guo2010}), while the purple cluster is enriched with genes responsible for the formation of inner cell mass (e.g., \textit{Pdgfra}, \textit{Pyy} \cite{Hou2007}). Reassuringly, these results of two underlying cell populations in the blastocyst are consistent with the known mammalian extra-embryonic lineages, the outer trophectoderm (TE) and internal primitive endoderm trophectoderm (PE), prior to implantation.


Notably, for both these single-cell data sets, most cells are assigned to a combination of more than one cluster, rather than a single cluster (the exception being the very early-stage cells in data from Deng et al). This highlights the potential utility for GoM models to capture structure in single cell data that might be missed by simpler cluster-based approaches.

The codes and scripts for reproducing results in this paper are available at \url{http://stephenslab.github.io/count-clustering/}.




%, 
%We find that admixture model is more successful in separating out different tissues in general, compared to the hierarchical clustering technique. The admixture model is essentially a count based modeling approach and seems to handle low counts and zero counts much better than the hierarchical method which is a more general approach to clustering. Since the RNA-seq data and in particular scRNA-seq data have lots of low counts and zero counts, the admixture model seems to be more suited for such data compared to hierarchical clustering method. 

%Currently there is a lot of interest in single cell sequencing as it is more informative about individual cell expression profiles compared to the RNA-seq on tissue samples. We were curious to see how stable the Admixture results are if the GTEx RNA-seq data is viewed at the scale of a single cell data. We achieve the latter by thinning the GTEx data under thinning parameter $p_{thin}=0.0001$ which is the order of scale obtained by dividing the total library size of the Jaitin \textit{et al} \cite{Jaitin2014} with respect to the library size of the GTEx V4 read counts data. We fitted the admixture model for $K=12$ on the thinned data and the Structure plot for the fitted model is presented in \textbf{Fig \ref{fig:fig4}}. It seems that most of the features observed in \textbf{Fig \ref{fig:fig1}} seem to be retained, for instance- the Brain samples clustering together, Whole blood and Testis forming separate clusters, Muscle skeletal and Heart tissue samples showing very similar patterns etc. However, thinning indeed shrinks the small differences across tissues and makes it more difficult to distinguish between tissues, as evident from the comparative study of hierarchical and admixture models, analogous to \textbf{Fig \ref{fig:fig3}}, for thinned data with thinning parameters $p_{thin}=0.001$ and $p_{thin}=0.0001$ in  \textbf{Fig \ref{fig:figS2}}. One can see that with thinning, the performance of admixture model in separating the tissues deteriorates but encouragingly, it seems that admixture does outperform the hierarchical clustering even under thinned data. \\[3pt]

 








