
\begin{thebibliography}{9}

%\bibitem{Sanders2014} 
%S Anders, T P Pyl, W Huber.
%\textit{HTSeq :  A Python framework to work with high-throughput sequencing data}. 
%Bioinformatics, 2014, in print; online at doi:10.1093/bioinformatics/btu638
%
%\bibitem{Li2009} 
%Li H.*, Handsaker B.*, Wysoker A., Fennell T., Ruan J., Homer N., Marth G., Abecasis G., Durbin R. and 1000 Genome Project Data Processing Subgroup.
%\textit{The Sequence alignment/map (SAM) format and SAMtools}. 
%Bioinformatics, 2009, 25, 2078-9. [PMID: 19505943]
%
%\bibitem{Liao2013}
%Liao Y, Smyth GK and Shi W. 
%\textit{The Subread aligner: fast, accurate and scalable read mapping by seed-and-vote}.
%Nucleic Acids Research, 2013, 41, pp. e108.

%\bibitem{Robinson2010}
%Robinson MD, McCarthy DJ and Smyth GK. 
%\textit{edgeR: a Bioconductor package for differential expression analysis of digital gene expression data}.
%Bioinformatics, 2010, 26, pp. -1.

%\bibitem{Ritchie2015}
%Ritchie ME, Phipson B, Wu D, Hu Y, Law CW, Shi W and Smyth GK. 
%\textit{limma powers differential expression analyses for RNA-sequencing and microarray studies}.
%Nucleic Acids Research, 2015, 43(7), pp. e47.

\bibitem{Jiang2011}
Jiang L, Schlesinger F, Davis CA, Zhang Y, Li R, Salit M, Gingeras TR and Oliver B. 
\textit{Synthetic spike-in standards for RNA-seq experiments}.
Genome Res, 2011:21, 1543-1551

\bibitem{Frazee2011}
Frazee AC, Langmead B, Leek JT. 
\textit{ReCount: a multi-experiment resource of analysis-ready RNA-seq gene count datasets}. 
BMC Bioinformatics, 2011, 12:449.

\bibitem{Zeisel2015}
Amit Zeisel, Ana B. Muñoz-Manchado, Simone Codeluppi, Peter Lönnerberg, Gioele La Manno, Anna Juréus, Sueli Marques, Hermany Munguba, Liqun He, Christer Betsholtz, Charlotte Rolny, Gonçalo Castelo-Branco, Jens Hjerling-Leffler, and Sten Linnarsson.
\textit{Cell types in the mouse cortex and hippocampus revealed by single-cell RNA-seq}.
Science 6 March 2015: 347 (6226), 1138-1142.

\bibitem{GTEX2013}
The GTEx Consortium. 
\textit{The Genotype-Tissue Expression (GTEx) project}. 
Nature genetics. 2013;45(6):580-585. doi:10.1038/ng.2653.

\bibitem{Taddy2012}
Matt Taddy.
\textit{On Estimation and Selection for Topic Models}. 
AISTATS 2012, JMLR W\&CP 22.
(maptpx R package).

\bibitem{Pritchard2000}
Pritchard, Jonathan K., Matthew Stephens, and Peter Donnelly. 
\textit{Inference of population structure using multilocus genotype data}. 
Genetics 155.2 (2000): 945-959.

\bibitem{Rosenberg2002}
Rosenberg NA, Pritchard JK, Weber JL, Cann HM, Kidd KK, Zhivotovsky LA and Feldman MW.
The genetic structure of human populations. 
Science 2002, 298: 2381-2385. 

\bibitem{Raj2014}
Anil Raj, Matthew Stephens, and Jonathan K. Pritchard. 
\textit{fastSTRUCTURE: Variational Inference of Population Structure in Large SNP Data Sets.}
Genetics.2014 197:573-589.

\bibitem{Maaten2008}
L.J.P. van der Maaten and G.E. Hinton. 
\textit{Visualizing High-Dimensional Data Using t-SNE}. 
Journal of Machine Learning Research, 2008: 2579-2605.

\bibitem{Maaten2014}
L.J.P. van der Maaten. 
\textit{Accelerating t-SNE using Tree-Based Algorithms}. 
Journal of Machine Learning Research, 2014:3221-3245.

\bibitem{Thompson2014}
Mark A, Thompson R and Wu C. 
\textit{mygene: Access MyGene.Info services}. 
2014. R package version 1.2.3.

%\bibitem{Law2014}
%Law CW, Chen Y, Shi W, Smyth GK. 
%\textit{voom: precision weights unlock linear model analysis tools for RNA-seq read counts}. 
%Genome Biology. 2014;15(2):R29. 

\bibitem{Jaitin2014}
Jaitin DA, Kenigsberg E et al.
\textit{Massively Parallel Single-Cell RNA-Seq for Marker-Free Decomposition of Tissues into Cell Types}.
Science, 2014: 343 (6172) 776-779.

\bibitem{Deng2014}
Deng Q, Ramskold D, Reinius B and Sandberg R.
\textit{Single-Cell RNA-Seq Reveals Dynamic, Random Monoallelic Gene Expression in Mammalian Cells}.
Science, 2014: 343 (6167) 193-196.


\bibitem{Leek2010}
Jeffrey T. Leek, Robert B. Scharpf, Héctor C Bravo, David Simcha, Benjamin Langmead, W. Evan Johnson, Donald Geman, Keith Baggerly and Rafael A. Irizarry
\textit{Tackling the widespread and critical impact of batch effects in high-throughput data}.
Nature Reviews Genetics 11, 733-739.

\bibitem{Hicks2015}
Stephanie C Hicks, Mingxiang Teng and Rafael A Irizarry
\textit{On the widespread and critical impact of systematic bias and batch effects in single-cell RNA-Seq data}.
BiorXiv, http://biorxiv.org/content/early/2015/09/04/025528

\bibitem{Houzel2005}
Herculano-Houzel S and Lent R.
\textit{Isotropic fractionator: a simple, rapid method for the quantification of total cell and neuron numbers in the brain}.
J Neurosci. 2005 Mar 9;25(10), 2518-21.

\bibitem{Thompson2014}
Mark A, Thompson R and Wu C. 
\textit{mygene: Access MyGene.Info services. R package version 1.2.3.}

\bibitem{Grün2015}
Grün D, Lyubimova A, Kester L, Wiebrands K, Basak O,  Sasaki N,  Clevers H, van Oudenaarden A.
\textit{Single-cell messenger RNA sequencing reveals rare intestinal cell types}.
Nature. 2015 Sep 10;525(7568), 251-5.

\bibitem{Buettner2015}
Buettner F, Natarajan KN, Casale FP, Proserpio V, Scialdone A, Theis FJ, Teichmann SA, Marioni JC and Stegle O.
\textit{Computational analysis of cell-to-cell heterogeneity in single-cell RNA-sequencing data reveals hidden subpopulations of cells}
Nature Biotechnology 2015, 33, 155?160, doi:10.1038/nbt.3102

\bibitem{Palmer2005}
Palmer C, Diehn M, Alizadeh AA and Brown PO.
\textit{Cell-type specific gene expression profiles of leukocytes in human peripheral blood}.
BMC Genomics 2006, 7:115.

\bibitem{Flutre2013}
Flutre T, Wen X, Pritchard J and Stephens M.
\textit{A Statistical Framework for Joint eQTL Analysis in Multiple Tissues}
PLoS Genet 2013, 9(5): e1003486. doi:10.1371/journal.pgen.1003486

\bibitem{Blei2003}
Blei DM, Ng AY  and Jordan MI.
\textit{Latent Dirichlet Allocation}
Journal of Machine Learning Research 2003, 3, 993-1022

\bibitem{Blei2009}
Blei DM and Lafferty J.
\textit{Topic Models}
In A. Srivastava and M. Sahami, editors, Text Mining: Classification, Clustering, and Applications . Chapman $\&$ Hall/CRC Data Mining and Knowledge Discovery Series, 2009.

\bibitem{Shen-Orr2010}
 Shen-Orr SS, Tibshirani R, Khatri, P, Bodian DL, Staedtler F, Perry NM,
 Hastie, T, Sarwal MM, Davis MM, Butte AJ.
 Cell typespecific gene expression   differences   in   complex   tissues.   
 Nature   Methods 2010,   7(4),   287?289  
 
\bibitem{Qiao2012}
Qiao W, Quon G,Csaszar E, Yu  M,  Morris Q, Zandstra  PW   
PERT: A   Method   for   Expression   Deconvolution   of   Human   Blood   Samples   from   Varied
 Microenvironmental   and   Developmental   Conditions.   
 PLoS   Comput   Biol  2012,  8(12)
 
 \bibitem{Repsilber2010}
 Repsilber D,  Kern   S,  Telaar   A,  Walzl   G,  Black   GF,   Selbig   J,   Parida  SK, Kaufmann SH, Jacobsen M.
 Biomarker discovery in heterogeneous tissue samples -taking the in-silico deconfounding approach. 
 BMC bioinformatics 2010,  11(1), 27+
 
 \bibitem{Schwartz2010}
 Schwartz R, Shackney SE. 
 Applying unmixing to gene expression data for tumor phylogeny inference. 
 BMC bioinformatics 2010,  11(1), 42+
 
 \bibitem{Lindsay2013}
 Lindsay J,  Mandoiu I, Nelson C.
 Gene Expression Deconvolution using Single-cells
 http://dna.engr.uconn.edu/bibtexmngr/upload/Lal.13.pdf.
 
\end{thebibliography}

